---
title: "Lecture 1: Asking Questions with Data"
subtitle: "Economic questions, data, and the probability toolkit"
author: "Swapnil Singh"
date: "2026-09-01"
date-format: "D MMMM YYYY"
footer: "Topics in Econometrics I · Lecture 1"
knitr:
  opts_chunk:
    out.width: "80%"
---

```{r}
#| label: setup
#| include: false

library(ggplot2)
library(patchwork)

set.seed(2026)

course_red <- "#E85D4A"
course_red_dark <- "#B83D2F"
course_charcoal <- "#242424"
course_grey <- "#DEDBD6"
course_pale <- "#FBE9E5"

theme_set(
  theme_minimal(base_size = 18) +
    theme(
      plot.title = element_text(
        colour = course_charcoal,
        face = "bold"
      ),
      plot.subtitle = element_text(colour = "#68645F"),
      panel.grid.minor = element_blank(),
      strip.text = element_text(face = "bold"),
      plot.background = element_rect(fill = "transparent", colour = NA)
    )
)

# Simulated school-district data: the data-generating process is known here.
n_districts <- 240
school_data <- data.frame(
  district = seq_len(n_districts),
  advantage = rnorm(n_districts)
)
school_data$class_size <- 23 - 2.2 * school_data$advantage +
  rnorm(n_districts, sd = 1.8)
school_data$test_score <- 665 - 0.7 * school_data$class_size +
  11 * school_data$advantage + rnorm(n_districts, sd = 5.5)
school_data$advantage_group <- cut(
  school_data$advantage,
  breaks = quantile(school_data$advantage, probs = c(0, 1 / 3, 2 / 3, 1)),
  include.lowest = TRUE,
  labels = c("Lower", "Middle", "Higher")
)

# A small illustrative panel.
panel_data <- expand.grid(
  year = 2019:2023,
  city = c("Aster", "Birch", "Cedar", "Dover", "Elm", "Flint")
)
city_effect <- c(Aster = 6, Birch = 2, Cedar = -2, Dover = 4, Elm = -5, Flint = 0)
panel_data$ridership <- 54 + city_effect[panel_data$city] +
  2.2 * (panel_data$year - 2019) + rnorm(nrow(panel_data), sd = 1.8)

# Objects used for probability illustrations.
failure_distribution <- data.frame(
  failures = 0:4,
  probability = c(0.80, 0.10, 0.06, 0.03, 0.01)
)

commute_joint <- data.frame(
  Weather = c("Rain", "No rain", "Total"),
  `Long commute` = c(0.15, 0.07, 0.22),
  `Short commute` = c(0.15, 0.63, 0.78),
  Total = c(0.30, 0.70, 1.00),
  check.names = FALSE
)
```

## Before we begin

::: {.key-idea}
**Econometrics is disciplined learning from imperfect data.**

We will keep asking: *What can the data tell us—and what must come from assumptions?*
:::

::: {.incremental}

- Bring a question, not merely a dataset.
- Translate words into a measurable target.
- Treat uncertainty as part of the answer.

:::

::: {.notes}
Timing: 2 minutes. Ask for one example of a quantitative economic question from the room. Keep the answer for the final slide.
:::

## By the end, you should be able to…

::: {.incremental}

1. distinguish **description**, **prediction**, and **causal inference**;
2. recognize cross-sectional, time-series, and panel data;
3. interpret a probability distribution, mean, variance, and conditional mean;
4. explain sampling variation, the law of large numbers, and the central limit theorem;
5. say what extra assumptions a claim needs.

:::

::: {.notes}
Timing: 2 minutes. Return to these outcomes in the exit ticket.
:::

## Part I · Economic questions {.section-slide background-color="#E85D4A"}

Good analysis begins before the first line of code.

::: {.notes}
Clock: 00:00–00:35, including the opening and a four-minute design activity.
:::

## Some questions

::: columns
::: {.column width="48%"}

### Education

How much would smaller classes improve achievement?

### Lending

Are otherwise similar applicants treated differently by race?

:::
::: {.column width="48%"}

### Public health

How much would a cigarette tax reduce smoking?

### Macroeconomy

How fast will GDP grow next year?

:::
:::

::: {.notes}
Timing: 2 minutes. These are the four motivating questions in Chapter 1. Ask which one is different in its inferential goal: GDP forecasting is predictive; the first three are causal.
:::

## Same data, different goals

::: columns
::: {.column width="31%"}

### Describe

What is the average test score among districts in our data?

**Summarize what we observed.**

:::
::: {.column width="31%"}

### Predict

For a new district with 22 students per class, what score should we expect?

**Use known characteristics to forecast an outcome.**

:::
::: {.column width="31%"}

### Explain a cause

If a district reduced class size, how much would its score change?

**Compare what would happen with and without the change.**

:::
:::

::: {.key-idea .fragment}
A model does not choose the goal. **The question does.**
:::

::: {.notes}
Timing: 3 minutes. Use the same class-size setting for all three goals. No notation is needed yet; potential outcomes arrive on the next causal slide.
:::

## The chain of an econometric answer

::: {.incremental}

**Question** → What decision or comparison matters?

**Estimand** → What population quantity would answer it?

**Data** → What was measured, on whom, and when?

**Assumptions** → Why does the calculation target the estimand?

**Estimate + uncertainty** → What did this sample teach us?

:::

::: {.notes}
Timing: 2 minutes. Say “estimand” slowly: the target, not the estimate. Use the distinction between a population average and a sample average.
:::

## A tempting pattern

```{r}
#| label: raw-class-size
#| echo: false
#| fig-width: 10
#| fig-height: 5.7
#| out-width: "78%"

ggplot(school_data, aes(class_size, test_score)) +
  geom_point(colour = course_red, alpha = 0.55, size = 2.4) +
  geom_smooth(
    method = "lm", se = FALSE,
    colour = course_charcoal, linewidth = 1.3
  ) +
  labs(
    title = "Districts with smaller classes have higher scores",
    subtitle = "Simulated cross-sectional data",
    x = "Students per class",
    y = "Average test score"
  )
```

::: {.notes}
Timing: 2 minutes. Ask: “Does this line tell us what would happen if a district hired teachers tomorrow?” Do not reveal the simulated data-generating process yet.
:::

## What comparison do we want?

For district $i$, imagine two outcomes:

$$
Y_i(1)=\text{score with a smaller class},\qquad
Y_i(0)=\text{score without it}.
$$

::: {.definition .fragment}
The individual causal effect is $Y_i(1)-Y_i(0)$. The average causal effect is

$$\operatorname{ATE}=E[Y_i(1)-Y_i(0)].$$
:::

::: {.key-idea .fragment}
**The missing counterfactual:** for the same district at the same moment, we observe only one of these two outcomes—never both.
:::

::: {.notes}
Timing: 3 minutes. Call this the missing counterfactual, not a missing-data problem that software can simply fill in.
:::

## Why random assignment is powerful

::: columns
::: {.column width="46%"}

### Treatment group

Randomly assigned to smaller classes

::: {.comparison-value}
**Average test score**
:::

::: {.comparison-caption}
smaller classes
:::

:::
::: {.column width="8%"}

###

**vs.**

:::
::: {.column width="46%"}

### Control group

Randomly assigned to usual classes

::: {.comparison-value}
**Average test score**
:::

::: {.comparison-caption}
usual classes
:::

:::
:::

::: {.key-idea .fragment}
With successful random assignment, the groups are comparable *before* treatment. The difference between their average scores estimates the causal effect.
:::

::: {.notes}
Timing: 2 minutes. Stress “in expectation” and “with a sufficiently large sample.” Randomization balances both observed and unobserved pre-treatment characteristics on average.
:::

## The ideal experiment is a benchmark

::: {.incremental}

- **Ethics:** We cannot randomly assign people to smoke.
- **Feasibility:** A multi-year school experiment is expensive and difficult.
- **Compliance:** Assigned treatment may differ from received treatment.
- **External validity:** Volunteers or study sites may not represent the target population.

:::

::: {.key-idea .fragment}
When an experiment is unavailable, ask: **What would the ideal experiment have compared?**
:::

::: {.notes}
Timing: 2 minutes. Connect to the Tennessee class-size experiment mentioned in the source chapter, without dwelling on historical details.
:::

## The third variable: district advantage

```{r}
#| label: confounding-plot
#| echo: false
#| fig-width: 10
#| fig-height: 5.7
#| out-width: "78%"

ggplot(school_data, aes(class_size, test_score, colour = advantage_group)) +
  geom_point(alpha = 0.50, size = 2.2) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.05) +
  scale_colour_manual(
    values = c("Lower" = course_red_dark, "Middle" = "#8B8580", "Higher" = course_charcoal),
    labels = c(
      "Lower" = "Lower advantage",
      "Middle" = "Middle advantage",
      "Higher" = "Higher advantage"
    ),
    name = "District advantage"
  ) +
  labs(
    title = "District advantage affects both variables",
    subtitle = "Higher family resources → smaller classes and higher scores (simulated)",
    x = "Students per class",
    y = "Average test score"
  )
```

::: {.notes}
Timing: 2 minutes. Because these data are simulated, we know the mechanism: district advantage affects class sizes and outcomes. In real data, the mechanism must be argued and investigated.
:::

## Association is not automatically causation

::: columns
::: {.column width="44%"}

$$
\text{District advantage}
\longrightarrow \text{Class size}
$$

$$
\text{District advantage}
\longrightarrow \text{Test score}
$$

:::
::: {.column width="52%"}

::: {.definition}
A **confounder** helps determine both the treatment and the outcome, creating a comparison between unlike units.
:::

:::
:::

::: {.key-idea .fragment}
Regression can adjust for measured differences. It cannot guarantee that every important difference was measured correctly.
:::

::: {.notes}
Timing: 2 minutes. This is a verbal causal diagram. Avoid claiming that “controlling for everything” solves identification.
:::

## Prediction does not require causation

::: {.exercise}
You see many umbrellas outside the library. Is “umbrella use” useful for predicting rain? Does opening an umbrella cause rain?
:::

::: {.answer .fragment}
Umbrellas can be highly predictive without being a cause of rain. Prediction needs a stable relationship; causal inference needs a credible comparison under intervention.
:::

::: {.notes}
Timing: 1 minute. This is the chapter’s core umbrella intuition. Elicit the answer before revealing it.
:::

## Active check · Design the comparison

::: {.exercise}
**Pairs · 4 minutes**

Question: *Does one additional hour of studying improve the next econometrics test score?*

1. State the treatment and outcome.
2. Describe an ideal randomized experiment.
3. Name one practical or ethical problem.
4. What would a purely predictive version of the question be?
:::

::: {.notes}
Timing: 4 minutes for pairs. Listen for treatment versions: “one hour” must be defined. Note spillovers and noncompliance if students raise them.
:::

## Design check · Debrief

::: {.answer}

- **Treatment:** a defined additional hour of study support before the test.
- **Outcome:** score on a specified test.
- **Ideal comparison:** randomly offer the support, then compare group means.
- **Complications:** compliance, spillovers, heterogeneous study quality, fairness.
- **Prediction:** “Given prior grades and study hours, what score should we expect?”

:::

::: {.key-idea .fragment}
The causal and predictive questions may use the same variables—and still require different logic.
:::

::: {.notes}
Timing: 1 minute for the debrief.
:::

## Part II · What kind of data? {.section-slide background-color="#B83D2F"}

Rows and columns have histories.

::: {.notes}
Clock: 00:35–00:55, including a short classification check.
:::

## Where did the data come from?

::: columns
::: {.column width="48%"}

### Experimental

Treatment is assigned by a designed mechanism.

**Strength:** comparability can come from randomization.

:::
::: {.column width="48%"}

### Observational

Behavior is recorded as it occurs through surveys, transactions, or administrative systems.

**Challenge:** treatment choices are usually not random.

:::
:::

::: {.small-note .fragment}
“Observational” does not mean “bad”; it means the design burden is different.
:::

::: {.notes}
Timing: 2 minutes. Name measurement error and selection as issues in either source; randomization does not fix a poorly measured outcome.
:::

## The shape of a dataset

| Data structure | Entities | Times | Main variation |
|---|---:|---:|---|
| Cross-section | many | one | across entities |
| Time series | one | many | across time |
| Panel / longitudinal | many | many | both dimensions |

::: {.key-idea .fragment}
The data structure determines which comparisons are available—and which uncertainty calculations are credible.
:::

::: {.notes}
Timing: 2 minutes. Introduce $n$ for entities and $T$ for periods. Mention repeated cross-sections as a fourth common structure, but do not develop it here.
:::

## Cross-section · many entities, one period

```{r}
#| label: cross-section-table
#| echo: false

school_snapshot <- school_data[c(7, 31, 88, 143, 219),
  c("district", "test_score", "class_size", "advantage_group")]
school_snapshot$test_score <- round(school_snapshot$test_score, 1)
school_snapshot$class_size <- round(school_snapshot$class_size, 1)
names(school_snapshot) <- c("District", "Test score", "Class size", "Advantage")

knitr::kable(school_snapshot, align = c("r", "r", "r", "l"))
```

::: {.definition}
Each row is a different district measured in the same period. Row order has no economic meaning.
:::

::: {.notes}
Timing: 2 minutes. Relate to the California district data in Chapter 1: $n=420$ districts in 1999.
:::

## Time series · one entity, many periods

```{r}
#| label: time-series-plot
#| echo: false
#| fig-width: 10
#| fig-height: 5.5

air_data <- data.frame(
  month = as.Date(paste0(time(AirPassengers), "-01"), format = "%Y.%m-%d"),
  passengers = as.numeric(AirPassengers)
)

ggplot(air_data, aes(month, passengers)) +
  geom_line(colour = course_charcoal, linewidth = 1) +
  geom_point(colour = course_red, size = 1.1, alpha = 0.65) +
  labs(
    title = "One system observed repeatedly",
    subtitle = "International airline passengers, 1949–1960 (built-in R series)",
    x = NULL,
    y = "Passengers (thousands)"
  )
```

::: {.notes}
Timing: 2 minutes. Point out trend, seasonality, and dependence across adjacent months. Time order matters, unlike row order in a cross-section.
:::

## Panel · many entities, many periods

```{r}
#| label: panel-plot
#| echo: false
#| fig-width: 10
#| fig-height: 5.4

ggplot(panel_data, aes(year, ridership, group = city, colour = city)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2.2) +
  scale_x_continuous(breaks = 2019:2023) +
  scale_colour_manual(
    values = c(course_red, course_charcoal, "#817C77", "#BC766B", "#4E4B48", "#D9A199")
  ) +
  labs(
    title = "Compare cities—and compare each city with itself",
    subtitle = "Illustrative public-transport ridership index",
    x = NULL,
    y = "Ridership index",
    colour = "City"
  )
```

::: {.notes}
Timing: 2 minutes. Relate to the chapter’s state-by-year cigarette data: $n=48$, $T=11$, and 528 state-year observations.
:::

## Active check · Name the structure

::: {.exercise}
Classify each dataset.

1. Wages of 2,000 workers surveyed in March 2026.
2. Lithuania’s quarterly GDP from 2000 to 2025.
3. Annual emissions for 27 EU countries from 1995 to 2025.
4. A new sample of households surveyed every year from 2010 to 2025.
:::

::: {.answer .fragment}
1. Cross-section · 2. Time series · 3. Panel · 4. Repeated cross-sections, unless the same households are followed.
:::

::: {.notes}
Timing: 3 minutes. The fourth item checks whether students track entities, not merely repeated dates.
:::

## Pause · 15 minutes {.section-slide background-color="#E85D4A"}

When we return: turning randomness into quantities.

::: {.notes}
Clock: 00:55–01:10. Leave this slide visible during the pause.
:::

## Part III · Probability distributions {.section-slide background-color="#E85D4A"}

First describe the possible values and how probability is distributed across them.

::: {.notes}
Clock: 01:10–01:50. This expanded review follows Chapter 2 from outcomes to probability distributions.
:::

## Why probability appears in econometrics

We observe one sample, but a different sample would give a different answer.

::: {.incremental}

- Which outcomes were possible?
- How frequently would each occur in repeated sampling?
- How much does an estimate vary across samples?
- Is the observed estimate surprising under a proposed model?

:::

::: {.key-idea .fragment}
Probability connects the dataset we have to the samples we might have drawn.
:::

::: {.notes}
Timing: 2 minutes. Distinguish empirical uncertainty from vagueness. The model describes a repeatable sampling thought experiment.
:::

## Random process → outcomes → sample space

Suppose a wireless connection is used while writing a term paper.

::: columns
::: {.column width="31%"}

### Random process

Use the connection for one writing session.

:::
::: {.column width="31%"}

### Outcome

The connection fails 0, 1, 2, 3, or 4 times.

Only one outcome occurs in a session.

:::
::: {.column width="31%"}

### Sample space

The set of all possible outcomes:

$$
\mathcal S=\{0,1,2,3,4\}.
$$

:::
:::

::: {.definition .fragment}
An **event** is a collection of outcomes. “At most one failure” is the event $\{0,1\}$.
:::

::: {.notes}
Timing: 3 minutes. Outcomes are mutually exclusive; the sample space is exhaustive. Events can contain one outcome or several outcomes.
:::

## A random variable turns outcomes into numbers

::: {.definition}
A **random variable** is a rule that assigns a numerical value to every possible outcome of a random process.
:::

| Random process | Random variable | Possible values |
|---|---|---|
| Meet a recent graduate | $G=1$ if employed, $0$ otherwise | $\{0,1\}$ |
| Use a wireless network | $M=$ number of failures | $\{0,1,2,\ldots\}$ |
| Commute to campus | $Y=$ time in minutes | positive real numbers |
| Observe a daily stock return | $R=$ percentage return | negative or positive real numbers |

::: {.key-idea .fragment}
“Random” means the value is unknown **before** the outcome occurs—not that the variable is measured carelessly.
:::

::: {.notes}
Timing: 3 minutes. Read one row from process to numerical variable. A realized value such as Y = 19.7 is no longer random after it is observed.
:::

## One outcome can generate several random variables

Consider one randomly selected commute to campus.

::: columns
::: {.column width="48%"}

### Commute time

$$Y=\text{minutes}$$

Continuous: 14.2, 19.7, 31.4, $\ldots$

:::
::: {.column width="48%"}

### Long commute

$$L=1\{Y>20\}$$

Binary: either $0$ or $1$.

:::
:::

::: {.key-idea .fragment}
The random process supplies the uncertainty. A random variable selects one numerical feature of its outcome.
:::

::: {.notes}
Timing: 2 minutes. The same commute yields a continuous variable $Y$ and a binary transformation $L$. Stress that $L$ is constructed from $Y$.
:::

## Discrete or continuous?

::: columns
::: {.column width="48%"}

### Discrete random variable

Takes a **countable** set of values.

- Employment indicator: $0$ or $1$
- Number of children: $0,1,2,\ldots$
- Number of loan defaults: $0,1,2,\ldots$

:::
::: {.column width="48%"}

### Continuous random variable

Can take any value in an **interval**.

- Earnings: €31,240.72
- Commute time: 19.73 minutes
- Inflation: $2.46\%$

:::
:::

::: {.key-idea .fragment}
The distinction determines how probabilities are represented: **bars at possible values** for discrete variables, **areas under curves** for continuous variables.
:::

::: {.notes}
Timing: 2 minutes. Recorded continuous variables are rounded in practice; the model can still treat the underlying quantity as continuous.
:::

## A discrete probability distribution

The **probability mass function** lists every possible value and its probability:

$$
p_M(m)=\Pr(M=m),\qquad p_M(m)\ge0,\qquad \sum_m p_M(m)=1.
$$

```{r}
#| label: failure-distribution-plot
#| echo: false
#| fig-width: 9.5
#| fig-height: 4.4
#| out-width: "68%"

ggplot(failure_distribution, aes(failures, probability)) +
  geom_col(fill = course_red, width = 0.68) +
  geom_text(
    aes(label = scales::percent(probability, accuracy = 1)),
    vjust = -0.35,
    colour = course_charcoal,
    size = 6
  ) +
  scale_x_continuous(breaks = 0:4) +
  scale_y_continuous(
    limits = c(0, 0.90),
    labels = scales::percent_format()
  ) +
  labs(
    title = "Possible network failures and their probabilities",
    x = "Number of failures, M",
    y = "Probability"
)
```

::: {.notes}
Timing: 3 minutes. Check that probabilities are nonnegative and sum to one. To obtain an event probability, sum the bars belonging to that event. This adapts the wireless-network example in Chapter 2.
:::

## From outcomes to events and cumulative probabilities

Using the network-failure distribution:

::: columns
::: {.column width="31%"}

### One outcome

$$
\Pr(M=0)=0.80
$$

No failures.

:::
::: {.column width="31%"}

### Several outcomes

$$
0.10+0.06=0.16
$$

One or two failures.

:::
::: {.column width="31%"}

### Cumulative event

$$
0.80+0.10=0.90
$$

At most one failure.

:::
:::

::: {.definition .fragment}
The **cumulative distribution function** is $F_M(m)=\Pr(M\le m)$: probability accumulated up to $m$.
:::

::: {.notes}
Timing: 3 minutes. Ask which bars are included before revealing each result. “At most” signals a cumulative probability.
:::

## Bernoulli distribution · yes or no

::: columns
::: {.column width="43%"}

### Definition

A Bernoulli variable records **one yes-or-no outcome**:

$$
D\sim\operatorname{Bernoulli}(p)
$$

$$
\Pr(D=1)=p,\qquad \Pr(D=0)=1-p.
$$

Examples: employed or not, default or repayment, treatment or control.

The support is always $\{0,1\}$; $p$ moves probability between the two bars.

:::
::: {.column width="53%"}

```{r}
#| label: bernoulli-distribution-plot
#| echo: false
#| fig-width: 6.2
#| fig-height: 4.7
#| out-width: "100%"

bernoulli_data <- data.frame(
  outcome = c(0, 1),
  probability = c(0.70, 0.30)
)

ggplot(bernoulli_data, aes(outcome, probability)) +
  geom_col(fill = course_red, width = 0.55) +
  geom_text(
    aes(label = scales::percent(probability)),
    vjust = -0.4,
    size = 6
  ) +
  scale_x_continuous(breaks = 0:1) +
  scale_y_continuous(
    limits = c(0, 0.80),
    labels = scales::percent_format()
  ) +
  labs(
    title = "Bernoulli(0.30)",
    x = "Outcome, D",
    y = "Probability"
  )
```

:::
:::

::: {.notes}
Timing: 3 minutes. In the plot, p = 0.30. Its mean and variance are derived after those concepts have been defined.
:::

## Two more discrete families

::: columns
::: {.column width="48%"}

### Binomial distribution

$B\sim\operatorname{Binomial}(n,p)$ counts successes in $n$ independent Bernoulli trials.

Example: defaults among 10 independent loans, each with default probability $0.30$.

:::
::: {.column width="48%"}

### Poisson distribution

$C\sim\operatorname{Poisson}(\lambda)$ counts events in a fixed interval with average rate $\lambda$.

Example: customer arrivals per hour when the average is three.

:::
:::

```{r}
#| label: discrete-family-plot
#| echo: false
#| fig-width: 10.5
#| fig-height: 3.7
#| out-width: "76%"

binomial_values <- 0:10
poisson_values <- 0:12
discrete_family_data <- rbind(
  data.frame(
    value = binomial_values,
    probability = dbinom(binomial_values, size = 10, prob = 0.30),
    family = "Binomial(10, 0.30)"
  ),
  data.frame(
    value = poisson_values,
    probability = dpois(poisson_values, lambda = 3),
    family = "Poisson(3)"
  )
)

ggplot(discrete_family_data, aes(value, probability)) +
  geom_col(fill = course_red, width = 0.72) +
  facet_wrap(~family, nrow = 1) +
  scale_x_continuous(breaks = 0:12) +
  labs(x = "Count", y = "Probability")
```

::: {.notes}
Timing: 3 minutes. This is a short extension beyond the book’s Bernoulli example. Emphasize that both are count distributions but arise from different stories.
:::

## Continuous probability distributions

A continuous variable can take any value in an interval. Its distribution is described by a **density**, $f_Y(y)$.

$$
f_Y(y)\ge0,\qquad
\int_{-\infty}^{\infty}f_Y(y)\,dy=1,\qquad
\Pr(a\le Y\le b)=\int_a^b f_Y(y)\,dy.
$$

::: columns
::: {.column width="48%"}

### Probability is area

The area under the curve between $a$ and $b$ is a probability.

:::
::: {.column width="48%"}

### A point has no area

For a continuous variable, $\Pr(Y=y)=0$ at any exact value.

:::
:::

::: {.key-idea .fragment}
A density’s **height is not itself a probability**. Probability is represented by area.
:::

::: {.notes}
Timing: 3 minutes. Contrast a density height with a probability. A density can exceed one as long as its total area remains one.
:::

## PDF and CDF · two views of one distribution

```{r}
#| label: pdf-cdf-commute-plot
#| echo: false
#| fig-width: 11.5
#| fig-height: 4.0
#| out-width: "84%"

commute_sdlog <- (log(20) - log(15)) / (qnorm(0.78) - qnorm(0.20))
commute_meanlog <- log(15) - commute_sdlog * qnorm(0.20)
commute_grid <- data.frame(minutes = seq(8, 28, length.out = 600))
commute_grid$density <- dlnorm(
  commute_grid$minutes,
  meanlog = commute_meanlog,
  sdlog = commute_sdlog
)
commute_grid$cdf <- plnorm(
  commute_grid$minutes,
  meanlog = commute_meanlog,
  sdlog = commute_sdlog
)

pdf_plot <- ggplot(commute_grid, aes(minutes, density)) +
  geom_area(
    data = subset(commute_grid, minutes >= 15 & minutes <= 20),
    fill = course_red,
    alpha = 0.55
  ) +
  geom_line(colour = course_charcoal, linewidth = 1.1) +
  labs(
    title = "Density (PDF)",
    subtitle = "Area from 15 to 20 minutes = 58%",
    x = "Commute time",
    y = "Density"
  )

cdf_plot <- ggplot(commute_grid, aes(minutes, cdf)) +
  geom_line(colour = course_red_dark, linewidth = 1.2) +
  geom_point(
    data = data.frame(minutes = c(15, 20), cdf = c(0.20, 0.78)),
    size = 3,
    colour = course_charcoal
  ) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Cumulative distribution (CDF)",
    subtitle = "20% by 15 minutes; 78% by 20 minutes",
    x = "Commute time",
    y = "Probability accumulated"
  )

pdf_plot + cdf_plot
```

::: {.key-idea .fragment}
The PDF shows **where probability is concentrated**; the CDF shows **how much probability has accumulated** by each value.
:::

::: {.notes}
Timing: 4 minutes. This reproduces the logic of Figure 2.2 in the book: 78% minus 20% gives 58% between 15 and 20 minutes.
:::

## Normal distribution · symmetric and continuous

::: columns
::: {.column width="40%"}

### Definition

$$
Y\sim N(\mu,\sigma^2)
$$

- Values can range from $-\infty$ to $\infty$.
- $\mu$ sets the center.
- $\sigma$ controls the spread.
- The curve is symmetric and bell-shaped.
- About 95% lies within $1.96\sigma$ of $\mu$.

Examples: measurement error, standardized test scores, and many sample averages.

:::
::: {.column width="56%"}

```{r}
#| label: normal-distribution-plot
#| echo: false
#| fig-width: 6.8
#| fig-height: 5.2
#| out-width: "100%"

normal_grid <- data.frame(score = seq(35, 105, length.out = 600))
normal_grid$density <- dnorm(normal_grid$score, mean = 70, sd = 10)
normal_grid$inside <- normal_grid$score >= 70 - 1.96 * 10 &
  normal_grid$score <= 70 + 1.96 * 10

ggplot(normal_grid, aes(score, density)) +
  geom_area(
    data = subset(normal_grid, inside),
    fill = course_red,
    alpha = 0.45
  ) +
  geom_line(colour = course_charcoal, linewidth = 1.2) +
  geom_vline(xintercept = 70, linetype = "dashed", colour = course_red_dark) +
  labs(
    title = "N(70, 10²)",
    subtitle = "95% lies between 50.4 and 89.6",
    x = "Test score",
    y = "Density"
  )
```

:::
:::

::: {.notes}
Timing: 4 minutes. The second parameter is the variance, not the standard deviation. The normal distribution has skewness zero and kurtosis three.
:::

## Log-normal distribution · positive and right-skewed

::: columns
::: {.column width="40%"}

### Definition

$$
\log(Y)\sim N(\mu,\sigma^2)
$$

- $Y$ is always positive.
- Most observations are moderate.
- A long right tail allows occasional large values.
- Taking logs makes the distribution symmetric.

Examples: earnings, wealth, firm size, prices, and durations.

Normality of $\log(Y)$ does **not** mean that $Y$ itself is normal.

:::
::: {.column width="56%"}

```{r}
#| label: lognormal-distribution-plot
#| echo: false
#| fig-width: 6.8
#| fig-height: 5.2
#| out-width: "100%"

lognormal_grid <- data.frame(
  minutes = seq(8, 36, length.out = 600)
)
lognormal_grid$density <- dlnorm(
  lognormal_grid$minutes,
  meanlog = commute_meanlog,
  sdlog = commute_sdlog
)
lognormal_grid$inside <- lognormal_grid$minutes >= 15 &
  lognormal_grid$minutes <= 20

ggplot(lognormal_grid, aes(minutes, density)) +
  geom_area(
    data = subset(lognormal_grid, inside),
    fill = course_red,
    alpha = 0.50
  ) +
  geom_line(colour = course_charcoal, linewidth = 1.2) +
  labs(
    title = "A positive, right-skewed shape",
    subtitle = "Shaded: 58% between 15 and 20 minutes",
    x = "Commute time (minutes)",
    y = "Density"
  )
```

:::
:::

::: {.notes}
Timing: 4 minutes. Earnings, prices, firm size, and durations are often right-skewed, although log-normality still must be checked rather than assumed.
:::

## Chi-squared distribution · sums of squares

::: columns
::: {.column width="40%"}

### Definition

If $Z_1,\ldots,Z_m$ are independent standard normal variables, then

$$
W=Z_1^2+\cdots+Z_m^2\sim\chi_m^2.
$$

- Values are nonnegative.
- The shape is right-skewed.
- $m$ is the **degrees of freedom**.
- Larger $m$ shifts the distribution right and makes it less skewed.

Later role: tests built from sums of squared standardized quantities.

:::
::: {.column width="56%"}

```{r}
#| label: chi-squared-plot
#| echo: false
#| fig-width: 6.8
#| fig-height: 5.2
#| out-width: "100%"

chi_x <- seq(0.05, 25, length.out = 700)
chi_data <- rbind(
  data.frame(x = chi_x, density = dchisq(chi_x, 2), df = "2 df"),
  data.frame(x = chi_x, density = dchisq(chi_x, 5), df = "5 df"),
  data.frame(x = chi_x, density = dchisq(chi_x, 10), df = "10 df")
)

ggplot(chi_data, aes(x, density, colour = df)) +
  geom_line(linewidth = 1.2) +
  scale_colour_manual(
    values = c("2 df" = course_red_dark, "5 df" = course_red, "10 df" = "#68645F")
  ) +
  labs(
    title = "Chi-squared shapes",
    x = "Nonnegative value",
    y = "Density",
    colour = NULL
  ) +
  theme(legend.position = "bottom")
```

:::
:::

::: {.notes}
Timing: 3 minutes. Squaring removes signs, so the support cannot be negative. Do not develop hypothesis tests yet.
:::

## Student $t$ distribution · a heavier-tailed bell

::: columns
::: {.column width="40%"}

### Definition

If $Z\sim N(0,1)$ and $W\sim\chi_m^2$ independently, then

$$
T=\frac{Z}{\sqrt{W/m}}\sim t_m.
$$

- Values range from $-\infty$ to $\infty$.
- It is symmetric around zero with heavier tails than the normal.
- As $m$ grows, $t_m$ approaches $N(0,1)$.

Later role: inference when uncertainty is estimated.

:::
::: {.column width="56%"}

```{r}
#| label: student-t-plot
#| echo: false
#| fig-width: 6.8
#| fig-height: 4.7
#| out-width: "100%"

t_x <- seq(-5, 5, length.out = 700)
t_data <- rbind(
  data.frame(x = t_x, density = dnorm(t_x), family = "Normal"),
  data.frame(x = t_x, density = dt(t_x, 3), family = "t, 3 df"),
  data.frame(x = t_x, density = dt(t_x, 10), family = "t, 10 df")
)

ggplot(t_data, aes(x, density, colour = family)) +
  geom_line(linewidth = 1.2) +
  scale_colour_manual(
    values = c("Normal" = course_charcoal, "t, 3 df" = course_red_dark, "t, 10 df" = course_red)
  ) +
  labs(
    title = "Lower df → heavier tails",
    x = "Value",
    y = "Density",
    colour = NULL
  ) +
  theme(legend.position = "bottom")
```

:::
:::

::: {.notes}
Timing: 3 minutes. Point to the shoulders and tails, not only the center. By roughly 30 degrees of freedom, the t distribution is close to normal.
:::

## $F$ distribution · a ratio of variances

::: columns
::: {.column width="40%"}

### Definition

If $W\sim\chi_m^2$ and $V\sim\chi_n^2$ independently, then

$$
F=\frac{W/m}{V/n}\sim F_{m,n}.
$$

- Values are nonnegative.
- The distribution is usually right-skewed.
- It has numerator degrees of freedom $m$ and denominator degrees of freedom $n$.

Later role: comparing explained and unexplained variation in joint tests.

:::
::: {.column width="56%"}

```{r}
#| label: f-distribution-plot
#| echo: false
#| fig-width: 6.8
#| fig-height: 5.2
#| out-width: "100%"

f_x <- seq(0.01, 5, length.out = 700)
f_data <- rbind(
  data.frame(x = f_x, density = df(f_x, 5, 10), family = "F(5, 10)"),
  data.frame(x = f_x, density = df(f_x, 10, 30), family = "F(10, 30)")
)

ggplot(f_data, aes(x, density, colour = family)) +
  geom_line(linewidth = 1.2) +
  scale_colour_manual(
    values = c("F(5, 10)" = course_red_dark, "F(10, 30)" = course_red)
  ) +
  labs(
    title = "Positive and right-skewed",
    x = "Ratio",
    y = "Density",
    colour = NULL
  ) +
  theme(legend.position = "bottom")
```

:::
:::

::: {.notes}
Timing: 3 minutes. Keep the definition conceptual: an F statistic compares two scaled sources of variation. The two degrees of freedom control the reference shape.
:::

## Active check · Match the variable to a family

::: {.exercise}

1. Whether one borrower defaults.
2. Number of defaults among 10 independent loans with a common default probability.
3. Number of customers arriving during one hour at a stable average rate.
4. A positive, strongly right-skewed measure of firm size whose logarithm is bell-shaped.
5. A symmetric test statistic with heavier tails because its standard error is estimated.

:::

::: {.answer .fragment}
1. Bernoulli · 2. Binomial · 3. Poisson · 4. Log-normal · 5. Student $t$
:::

::: {.key-idea .fragment}
Choose a distribution by asking about the **support**, the **data-generating story**, and the **shape**—not by choosing the prettiest curve.
:::

::: {.notes}
Timing: 3 minutes. Have students answer in pairs. Ask what empirical evidence could make them reject each proposed family.
:::

## Part IV · Moments and relationships {.section-slide background-color="#B83D2F"}

Now summarize a distribution’s center, spread, shape, and co-movement.

::: {.notes}
Clock: 01:50–02:10. Keep derivations brisk; Part III now carries more of the probability review.
:::

## Mean · the center of a distribution

The **mean**, **expected value**, and **expectation** are the same population quantity, denoted $\mu_Y=E(Y)$.

::: columns
::: {.column width="48%"}

### Discrete $Y$

$$
E(Y)=\sum_y y\Pr(Y=y)
$$

:::
::: {.column width="48%"}

### Continuous $Y$

$$
E(Y)=\int_{-\infty}^{\infty} y f_Y(y)\,dy
$$

:::
:::

::: {.definition .fragment}
The mean is the **probability-weighted average** of all possible values—the long-run average across repeated draws.
:::

::: {.notes}
Timing: 3 minutes. This is the first moment. Stress that a mean need not be a possible realization and that both formulas express the same idea.
:::

## Mean

For network failures $M$,

$$
E(M)=\sum_m m\Pr(M=m)
$$

$$
=0(0.80)+1(0.10)+2(0.06)+3(0.03)+4(0.01)=\mathbf{0.35}.
$$

::: {.key-idea .fragment}
$E(M)=0.35$ does **not** predict 0.35 failures in one session. It is the average over many sessions.
:::

::: {.notes}
Timing: 2 minutes. Ask whether 0.35 must be a possible realization. For a Bernoulli variable, use the same formula to derive $E(D)=p$.
:::

## Variance · how far outcomes wander

$$
\operatorname{Var}(Y)=E[(Y-\mu_Y)^2],
\qquad
\operatorname{SD}(Y)=\sqrt{\operatorname{Var}(Y)}.
$$

::: columns
::: {.column width="48%"}

### Variance

Squares deviations, so large misses matter more.

Units are squared.

:::
::: {.column width="48%"}

### Standard deviation

Returns to the original units of $Y$.

Think “typical scale of variation,” not a guaranteed distance.

:::
:::

::: {.key-idea .fragment}
For a Bernoulli variable, $E(D)=p$ and $\operatorname{Var}(D)=p(1-p)$.
:::

::: {.notes}
Timing: 2 minutes. For the network example, variance is about 0.648 and SD about 0.80 failures.
:::

## What happens when units change?

If $Z=a+bY$, then

$$
E(Z)=a+bE(Y),
$$

$$
\operatorname{Var}(Z)=b^2\operatorname{Var}(Y),
\qquad
\operatorname{SD}(Z)=|b|\operatorname{SD}(Y).
$$

::: {.exercise .fragment}
If temperature in Celsius has mean $20$ and SD $4$, what are the mean and SD in Fahrenheit, where $F=32+1.8C$?
:::

::: {.answer .fragment}
$E(F)=68$ and $\operatorname{SD}(F)=7.2$. Adding 32 moves the center but does not change the spread.
:::

::: {.notes}
Timing: 3 minutes. Pause for mental arithmetic before revealing. Stress the absolute value for a negative scale factor.
:::

## Skewness · asymmetry

$$
\operatorname{Skew}(Y)=
\frac{E[(Y-\mu_Y)^3]}{\sigma_Y^3}.
$$

::: columns
::: {.column width="31%"}

### Negative

Longer left tail

:::
::: {.column width="31%"}

### Zero

Symmetric distribution

:::
::: {.column width="31%"}

### Positive

Longer right tail

:::
:::

::: {.key-idea .fragment}
Cubing preserves the sign of deviations; dividing by $\sigma_Y^3$ removes the units.
:::

::: {.notes}
Timing: 2 minutes. A symmetric distribution has skewness zero, but zero skewness alone does not guarantee full symmetry.
:::

## Kurtosis · tail weight

$$
\operatorname{Kurt}(Y)=
\frac{E[(Y-\mu_Y)^4]}{\sigma_Y^4}.
$$

::: {.incremental}

- Fourth powers give extreme deviations substantial weight.
- Larger kurtosis means more probability in the tails and greater outlier risk.
- The normal distribution has kurtosis $3$.
- A distribution with kurtosis above $3$ is **heavy-tailed** relative to the normal.

:::

::: {.notes}
Timing: 2 minutes. Use the book’s Wall Street example: a 21-standard-deviation event is essentially impossible under a constant-variance normal model, revealing much heavier empirical tails.
:::

## Same center and spread, different shapes

```{r}
#| label: distribution-shapes-plot
#| echo: false
#| fig-width: 11
#| fig-height: 4.9
#| out-width: "82%"

standardize_draws <- function(values) {
  as.numeric((values - mean(values)) / sd(values))
}

shape_data <- rbind(
  data.frame(
    value = standardize_draws(rnorm(12000)),
    family = "Normal\nsymmetric; kurtosis 3"
  ),
  data.frame(
    value = standardize_draws(rlnorm(12000, sdlog = 0.55)),
    family = "Log-normal\nright-skewed"
  ),
  data.frame(
    value = standardize_draws(rt(12000, df = 5)),
    family = "Student t₅\nsymmetric; heavy tails"
  )
)
shape_data$family <- factor(
  shape_data$family,
  levels = c(
    "Normal\nsymmetric; kurtosis 3",
    "Log-normal\nright-skewed",
    "Student t₅\nsymmetric; heavy tails"
  )
)

ggplot(shape_data, aes(value)) +
  geom_density(fill = course_pale, colour = course_red_dark, linewidth = 0.9) +
  facet_wrap(~family, nrow = 1) +
  coord_cartesian(xlim = c(-4, 6), ylim = c(0, 0.55)) +
  labs(
    title = "All three are standardized to mean 0 and standard deviation 1",
    x = "Standardized value",
    y = "Density"
  )
```

::: {.notes}
Timing: 3 minutes. Equal means and variances do not imply equal distributions. Connect the right-skewed panel to earnings and the heavy-tailed panel to financial returns.
:::

## Standardization gives a common scale

$$
Z=\frac{Y-\mu_Y}{\sigma_Y}
$$

::: columns
::: {.column width="48%"}

### Center

$E(Z)=0$

:::
::: {.column width="48%"}

### Scale

$\operatorname{Var}(Z)=1$

:::
:::

::: {.definition .fragment}
A $z$-score says how many standard deviations an observation lies above or below the mean.
:::

::: {.notes}
Timing: 2 minutes. Standardization changes units, not rank. It will reappear in the CLT and test statistics.
:::

## Moments · a compact map

| Quantity | Formula | What it describes |
|---|---|---|
| Mean | $E(Y)$ | Center |
| Variance | $E[(Y-\mu_Y)^2]$ | Spread |
| Skewness | $E[(Y-\mu_Y)^3]/\sigma_Y^3$ | Asymmetry |
| Kurtosis | $E[(Y-\mu_Y)^4]/\sigma_Y^4$ | Tail weight |

::: {.key-idea .fragment}
Mean and variance do not fully describe a distribution; shape and tail behavior can matter economically.
:::

::: {.notes}
Timing: 2 minutes. The raw $r$th moment is $E(Y^r)$. Variance, skewness, and kurtosis use centered or standardized moments.
:::

## From one variable to two {.section-slide background-color="#B83D2F"}

Joint and conditional distributions describe economic relationships.

::: {.notes}
Continue Part IV. The mean and variance are now defined, so conditional means and covariance have a clear foundation.
:::

## Joint and marginal distributions

```{r}
#| label: commute-joint-table
#| echo: false

commute_display <- commute_joint
for (column in 2:4) {
  commute_display[[column]] <- sprintf("%.2f", commute_display[[column]])
}
knitr::kable(commute_display, align = c("l", "r", "r", "r"))
```

::: {.incremental}

- Interior cells are **joint probabilities**.
- Row and column totals are **marginal probabilities**.
- All four joint probabilities sum to one.

:::

::: {.notes}
Timing: 2 minutes. This adapts the rainy-commute table in Chapter 2. Ask students to locate Pr(rain and long commute) and Pr(long commute).
:::

## Conditioning changes the reference group

What is the probability of a long commute **given rain**?

$$
\Pr(\text{long}\mid\text{rain})
=\frac{\Pr(\text{long and rain})}{\Pr(\text{rain})}
=\frac{0.15}{0.30}=0.50.
$$

::: {.definition .fragment}
Conditioning restricts attention to cases where the condition is true, then renormalizes probabilities within that group.
:::

::: {.notes}
Timing: 3 minutes. The denominator is the new reference population. Contrast Pr(long | rain) with Pr(rain | long).
:::

## Independence is a strong statement

$X$ and $Y$ are independent when learning $X$ does not change the distribution of $Y$:

$$
\Pr(Y=y\mid X=x)=\Pr(Y=y)
$$

or, equivalently,

$$
\Pr(X=x,Y=y)=\Pr(X=x)\Pr(Y=y).
$$

::: {.exercise .fragment}
Are rain and commute length independent in the table?
:::

::: {.answer .fragment}
No. $\Pr(\text{long}\mid\text{rain})=0.50$, while $\Pr(\text{long})=0.22$.
:::

::: {.notes}
Timing: 2 minutes. Independence concerns the full distribution, not merely a zero correlation.
:::

## Conditional mean · center within a group

$$
m(x)=E(Y\mid X=x)
$$

::: {.incremental}

- It is the mean of $Y$ among cases with $X=x$.
- It summarizes how the center of $Y$ changes with information $X$.
- It can vary with $x$, tracing a **conditional mean function**.

:::

::: {.key-idea .fragment}
Regression is, at its core, a way to model a conditional mean.
:::

::: {.notes}
Timing: 2 minutes. This prepares the language for later OLS. Mention in delivery, but do not prove, that the conditional mean minimizes mean squared prediction error.
:::

## The law of iterated expectations

::: {.definition}
For any random variables $Y$ and $X$,

$$
E(Y)=E\!\left[E(Y\mid X)\right].
$$

First find the mean of $Y$ within each value of $X$; then average those conditional means over $X$.
:::

### Proof for discrete $X$ and $Y$

$$
\begin{aligned}
E\!\left[E(Y\mid X)\right]
&= \sum_x E(Y\mid X=x)\Pr(X=x) \\
&= \sum_x\sum_y y\Pr(Y=y\mid X=x)\Pr(X=x) \\
&= \sum_y y\sum_x \Pr(Y=y,X=x) \\
&= \sum_y y\Pr(Y=y) \\
&= E(Y).
\end{aligned}
$$

::: {.notes}
Timing: 3 minutes. The second line expands the conditional mean. The third uses the multiplication rule, and the fourth sums out $X$. For continuous variables, the same argument replaces sums with integrals.
:::

## Covariance · unscaled co-movement

$$
\operatorname{Cov}(X,Y)=E[(X-\mu_X)(Y-\mu_Y)]
$$

::: columns
::: {.column width="48%"}

**Positive:** above-average $X$ tends to accompany above-average $Y$.

:::
::: {.column width="48%"}

**Negative:** above-average $X$ tends to accompany below-average $Y$.

:::
:::

::: {.definition .fragment}
Covariance is positive when $X$ and $Y$ tend to deviate from their means in the same direction. Its units are the units of $X$ multiplied by the units of $Y$.
:::

::: {.notes}
Timing: 2 minutes. Use the signs of the centered products for intuition.
:::

## Correlation · unit-free linear co-movement

$$
\operatorname{Corr}(X,Y)=
\frac{\operatorname{Cov}(X,Y)}{\sigma_X\sigma_Y},
\qquad -1\le\operatorname{Corr}(X,Y)\le1.
$$

::: {.incremental}

- Dividing by both standard deviations removes the units.
- The sign gives the direction of **linear** co-movement.
- The magnitude describes the strength of that linear pattern.
- Correlation does not by itself identify a causal effect.

:::

::: {.notes}
Timing: 2 minutes. Perfect correlation of plus or minus one means all probability lies on a line with positive or negative slope; zero correlation need not mean independence.
:::

## Reproduce it in R

```{r}
#| label: nonlinear-data

set.seed(2026)

n <- 500
x <- runif(n, -2, 2)
y <- x^2 + rnorm(n, sd = 0.35)

round(cor(x, y), 3)
```

::: {.key-idea .fragment}
**Yes.** Pearson correlation measures **linear** co-movement. Here $Y=X^2+\text{noise}$ is U-shaped, so its negative-$X$ and positive-$X$ sides offset in the covariance.
:::

::: {.notes}
Timing: 2 minutes. Run the code live if useful. For a symmetric $X$, $\operatorname{Cov}(X,X^2)=E(X^3)-E(X)E(X^2)=0$. Independent mean-zero noise does not change that population covariance.
:::

## Zero correlation can hide a relationship

```{r}
#| label: nonlinear-correlation
#| echo: false
#| fig-width: 9.8
#| fig-height: 5.4

nonlinear_data <- data.frame(
  x = x,
  y = y
)
nonlinear_correlation <- cor(nonlinear_data$x, nonlinear_data$y)

ggplot(nonlinear_data, aes(x, y)) +
  geom_point(colour = course_red, alpha = 0.45, size = 2) +
  geom_smooth(method = "loess", se = FALSE, colour = course_charcoal, linewidth = 1.2) +
  labs(
    title = "A strong relationship with almost zero correlation",
    subtitle = paste0("Simulated correlation = ", round(nonlinear_correlation, 2)),
    x = "X",
    y = "Y = X² + noise"
  )
```

::: {.notes}
Timing: 2 minutes. Correlation captures linear co-movement. Independence implies zero correlation under finite moments, but zero correlation does not imply independence.
:::

## Active check · Conditional or causal?

::: {.exercise}
A wage dataset shows

$$E(\text{wage}\mid\text{degree}=1) - E(\text{wage}\mid\text{degree}=0)=€6.40/\text{hour}.$$

1. one sentence this number **does** justify;
2. one causal sentence it **does not yet** justify;
3. two variables that might make the groups unlike.
:::

::: {.answer .fragment}
**Answers**

1. **Descriptive:** In this dataset, degree holders earn €6.40 more per hour on average.
2. **Not justified:** Obtaining a degree increases a person’s wage by €6.40 per hour.
3. **Possible group differences:** prior ability and family resources.
:::

::: {.notes}
Timing: 3 minutes plus a 1-minute debrief. Other candidate confounders include field, labor-market experience, and location. Avoid suggesting that adding a long control list automatically identifies the degree effect.
:::

## Part V · From samples to populations {.section-slide background-color="#E85D4A"}

The estimate changes when the sample changes.

::: {.notes}
Clock: 02:10–02:30. Emphasize the LLN definition and use the R simulation to connect the theorem to a visible sample path.
:::

## Population, sample, estimator, estimate

::: columns
::: {.column width="48%"}

### Population target

$$\mu_Y=E(Y)$$

Fixed but usually unknown.

:::
::: {.column width="48%"}

### Sample statistic

$$\overline{\,Y\,}=\frac{1}{n}\sum_{i=1}^{n}Y_i$$

Random before sampling; a number after the data arrive.

:::
:::

::: {.definition .fragment}
An **estimator** is the rule ($\overline{\,Y\,}$). An **estimate** is the realized value from this sample.
:::

::: {.notes}
Timing: 2 minutes. This distinction is foundational for standard errors and confidence intervals.
:::

## Simple random sampling and i.i.d.

We often begin with

$$Y_1,\ldots,Y_n\ \text{independent and identically distributed}.$$

::: {.incremental}

- **Identical distribution:** each observation comes from the same population distribution.
- **Independent:** learning one observation gives no information about another.

:::

::: {.small-note .fragment}
Time series, clustered students, households, firms, and spatial data often violate simple independence. Later methods adapt the uncertainty calculation.
:::

::: {.notes}
Timing: 2 minutes. Say explicitly that i.i.d. is a starting model, not a universal property of data.
:::

## The sample mean has a distribution

Imagine repeatedly drawing samples of $n$ students and recording average commute time.

::: {.incremental}

- Each sample contains different students.
- Each sample produces a different $\overline{\,Y\,}$.
- The distribution of those possible averages is the **sampling distribution**.

:::

::: {.key-idea .fragment}
A histogram of the observed $Y_i$ values is not the same object as the sampling distribution of $\overline{\,Y\,}$.
:::

::: {.notes}
Timing: 2 minutes. The population distribution describes individuals; the sampling distribution describes an estimator across repeated samples.
:::

## What averaging buys us

If $Y_1,\ldots,Y_n$ are i.i.d. with mean $\mu_Y$ and variance $\sigma_Y^2$, then

$$
E(\overline{\,Y\,})=\mu_Y,
\qquad
\operatorname{Var}(\overline{\,Y\,})=\frac{\sigma_Y^2}{n},
\qquad
\operatorname{SD}(\overline{\,Y\,})=\frac{\sigma_Y}{\sqrt n}.
$$

::: {.key-idea .fragment}
Four times the sample size gives half the standard deviation—not one quarter.
:::

::: {.notes}
Timing: 3 minutes. The first equality is unbiasedness. The last quantity motivates the standard error once sigma is estimated.
:::

## Law of large numbers · definition

Suppose $Y_1,Y_2,\ldots$ are i.i.d. with finite mean $E(Y_i)=\mu_Y$. Define

$$
\overline{\,Y\,}_n=\frac{1}{n}\sum_{i=1}^{n}Y_i.
$$

The **weak law of large numbers** states that

$$
\overline{\,Y\,}_n\xrightarrow{p}\mu_Y.
$$

::: {.definition .fragment}
Equivalently, for every $\varepsilon>0$,

$$
\lim_{n\to\infty}\Pr\!\left(\left|\overline{\,Y\,}_n-\mu_Y\right|>\varepsilon\right)=0.
$$
:::

::: {.notes}
Timing: 3 minutes. Read the probability statement in words: for any fixed tolerance, the probability that the sample mean misses the population mean by more than that tolerance goes to zero.
:::

## Law of large numbers · R code

```{r}
#| label: lln-code

set.seed(2026)
number_of_draws <- 500
lln_draws <- rexp(number_of_draws, rate = 1)
lln_data <- data.frame(
  n = seq_len(number_of_draws),
  running_mean = cumsum(lln_draws) / seq_len(number_of_draws)
)

lln_plot <- ggplot(lln_data, aes(n, running_mean)) +
  geom_hline(yintercept = 1, colour = course_red) +
  geom_line(colour = course_charcoal) +
  labs(x = "Sample size, n", y = "Running sample mean")
```

::: {.notes}
Timing: 2 minutes. The exponential population has mean one. Explain that `cumsum(lln_draws) / seq_len(number_of_draws)` computes every running average from the same sequence.
:::

## Law of large numbers · destination

```{r}
#| label: lln-plot
#| echo: false
#| fig-width: 10
#| fig-height: 5.5

lln_plot +
  annotate(
    "text", x = 440, y = 1.10,
    label = "Population mean = 1",
    colour = course_red_dark, size = 5.5
  ) +
  labs(
    title = "As the sample grows, the running average settles",
    subtitle = "One simulated sequence from a right-skewed exponential population"
  )
```

::: {.notes}
Timing: 2 minutes. One path illustrates, but does not prove, the LLN. The theorem concerns the probability of a large error across possible samples, not monotone convergence along every realized path.
:::

## Central limit theorem · shape

For large $n$,

$$
\frac{\overline{\,Y\,}-\mu_Y}{\sigma_Y/\sqrt n}
\overset{d}{\approx}N(0,1).
$$

::: {.incremental}

- The population itself need not be normal.
- A skewed or discrete population can produce nearly normal averages.
- “Large enough” depends on skewness, tails, and dependence.

:::

::: {.notes}
Timing: 2 minutes. Keep convergence notation informal. The standardized average has a stable scale; the unstandardized sampling distribution collapses toward mu.
:::

## Watching the CLT work

```{r}
#| label: clt-simulation
#| echo: false
#| fig-width: 11
#| fig-height: 5.6

sample_sizes <- c(1, 5, 30)
repetitions <- 5000

clt_data <- do.call(rbind, lapply(sample_sizes, function(sample_size) {
  sample_means <- replicate(
    repetitions,
    mean(rexp(sample_size, rate = 1))
  )
  data.frame(
    z = sqrt(sample_size) * (sample_means - 1),
    sample_size = factor(
      paste0("n = ", sample_size),
      levels = paste0("n = ", sample_sizes)
    )
  )
}))

normal_reference <- expand.grid(
  z = seq(-4, 5, length.out = 500),
  sample_size = factor(
    paste0("n = ", sample_sizes),
    levels = paste0("n = ", sample_sizes)
  )
)
normal_reference$density <- dnorm(normal_reference$z)

ggplot(clt_data, aes(z)) +
  geom_density(fill = course_pale, colour = course_red_dark, linewidth = 0.9) +
  geom_line(
    data = normal_reference,
    aes(z, density),
    colour = course_charcoal,
    linewidth = 1,
    inherit.aes = FALSE
  ) +
  facet_wrap(~sample_size, nrow = 1) +
  coord_cartesian(xlim = c(-4, 5), ylim = c(0, 0.48)) +
  labs(
    title = "Averages of skewed observations become approximately normal",
    subtitle = "Red density: simulated standardized averages · Black curve: N(0,1)",
    x = "Standardized sample average",
    y = "Density"
  )
```

::: {.notes}
Timing: 3 minutes. At n=1 the standardized variable retains exponential skewness. By n=30 the approximation is useful, though not magically exact.
:::

## LLN and CLT answer different questions

| Result | What it tells us | Mental image |
|---|---|---|
| Law of large numbers | $\overline{\,Y\,}$ gets close to $\mu_Y$ | the target becomes accurate |
| Central limit theorem | the standardized error becomes approximately normal | the error gets a useful shape |

::: {.key-idea .fragment}
LLN supports consistency. CLT supports approximate probabilities, confidence intervals, and tests.
:::

::: {.notes}
Timing: 2 minutes. This distinction is often blurred. The LLN is about concentration; the CLT is about distributional approximation.
:::

## Takeaways {.section-slide background-color="#E85D4A"}

Econometrics makes the chain visible:

**question → estimand → data → assumptions → evidence → decision**

::: {.notes}
Timing: 2 minutes. Revisit the opening student question and walk it through the chain in one minute.
:::
