---
title: "Lecture 3: Inference for Regression Coefficients"
subtitle: "Standard errors, hypothesis tests, and confidence intervals"
author: "Swapnil Singh"
date: "2026-09-22"
date-format: "D MMMM YYYY"
footer: "Topics in Econometrics I · Lecture 3"
format:
  revealjs:
    incremental: false
---

```{r}
#| label: setup
#| include: false

library(ggplot2)
```

## From a sampling distribution to a decision

Last lecture: the estimated slope $\hat\beta_1$ varies across samples.

::: {.key-idea}
**Today:** Which regression coefficient values are consistent with our sample?
:::

- A **standard error** estimates how much $\hat\beta_1$ varies across samples.
- A **$t$-statistic** measures the gap from a proposed value in standard-error units.
- A **$p$-value** turns that comparison into a test.
- A **confidence interval** collects coefficient values the test would retain.

::: {.notes}
Source: Stock and Watson, Chapter 5, Sections 5.1–5.2, supplied in
book/chapter5. Section 5.3 on binary regressors is beyond this deck.
:::

## Testing one regression coefficient {.section-slide background-color="#E85D4A"}

Section 5.1 · From a claim about a coefficient to evidence from a sample

::: {.notes}
The California estimates below are the textbook's reported results, not a
new analysis of district-level data.
:::

## A claim about class size

The textbook studies test scores and the student–teacher ratio in **420 California school districts**.

$$
\text{TestScore}_i=\beta_0+\beta_1\text{STR}_i+u_i.
$$

- A taxpayer claims that changing class size has no effect: $\beta_1=0$.
- The estimated slope is $\hat\beta_1=-2.28$ score points per additional student per teacher.
- Could a slope this far from zero arise through sampling variation?

::: {.notes}
STR is students per teacher, used as a measure of class size. The coefficient
is an association unless the regression's causal assumptions hold. A test
cannot establish those assumptions.
:::

## State the claim before examining the test

$$
H_0:\beta_1=\beta_{1,0}
\qquad\text{versus}\qquad
H_1:\beta_1\ne\beta_{1,0}.
$$

- $H_0$ is the **null hypothesis**: a specific population slope.
- $H_1$ is **two-sided**: the slope could lie on either side.
- For the taxpayer's claim, $\beta_{1,0}=0$.
- Choose a significance level, such as $\alpha=0.05$, **before** looking at the result.

::: {.notes}
The parameter is fixed and unknown; the estimate is random before sampling.
Failure to reject does not prove that the null is true.
:::

## The familiar $t$-statistic has a new estimator

$$
t=\frac{\text{estimator}-\text{hypothesized value}}
        {\text{standard error of the estimator}}.
$$

::: columns
::: {.column width="48%"}

### Population mean

$$t=\frac{\bar Y-\mu_{Y,0}}{SE(\bar Y)}$$

:::
::: {.column width="48%"}

### Regression slope

$$t=\frac{\hat\beta_1-\beta_{1,0}}{SE(\hat\beta_1)}$$

:::
:::

::: {.key-idea}
The numerator is a distance from the null; the denominator puts that distance on the scale of sampling uncertainty.
:::

::: {.notes}
Equations 5.1 and 5.5. Section 5.1 uses a large-sample standard normal
approximation, not an exact Student t distribution.
:::

## What the slope's standard error measures

::: {.definition}
$SE(\hat\beta_1)$ estimates the standard deviation of the **sampling distribution** of $\hat\beta_1$.
:::

- Imagine repeatedly drawing districts and fitting the same regression.
- The slope estimates would differ across samples.
- A larger standard error means the observed slope is less precise.
- It measures uncertainty about the estimated **slope**, not the spread of individual test scores.

::: {.notes}
The standard error differs from SER, which summarizes residual scale around
the fitted line.
:::

## Start with the OLS slope error

Recall the population model $Y_i=\beta_0+\beta_1X_i+u_i$. Write
$d_i=X_i-\bar X$ and $S_{XX}=\sum_i d_i^2$.

$$
\hat\beta_1
=\frac{\sum_i d_iY_i}{S_{XX}}
=\beta_1+\frac{\sum_i d_i u_i}{S_{XX}}.
$$

::: {.key-idea}
The fitted slope differs from the population slope because the sampled $X_i$ and unobserved $u_i$ happen to move together.
:::

- Larger spread in $X$ increases $S_{XX}$ and helps pin down the slope.
- This identity holds for the observed sample; no normality assumption is needed.

::: {.notes}
Recall Lecture 2's exact slope-error identity. Substitution gives
sum_i d_i beta_0=0 and sum_i d_i beta_1 X_i=beta_1 S_XX.
The distinction between the exact identity and the following variance
calculation matters: the next slide conditions on the observed X values.
:::

## Derive the slope's conditional variance

Write $X=(X_1,\ldots,X_n)$. Once we condition on $X$, the weights $d_i$ and denominator $S_{XX}$ are fixed:

$$
\operatorname{Var}(\hat\beta_1\mid X)
=\frac{1}{S_{XX}^2}\operatorname{Var}\!\left(\sum_i d_i u_i\mid X\right).
$$

For a weighted sum, add the individual variances and the pairwise covariances:

$$
\operatorname{Var}\!\left(\sum_i d_i u_i\mid X\right)
=\sum_i d_i^2\operatorname{Var}(u_i\mid X_i)
+\underbrace{2\sum_{i<j}d_i d_j\operatorname{Cov}(u_i,u_j\mid X)}_{=\,0\text{ with independent observations}}.
$$

::: {.key-idea}
Therefore $\displaystyle \operatorname{Var}(\hat\beta_1\mid X)
=\frac{\sum_i d_i^2\operatorname{Var}(u_i\mid X_i)}{S_{XX}^2}$.
:::

::: {.notes}
From the previous slide, beta-hat_1 = beta_1 + sum_i d_i u_i/S_XX.
A constant beta_1 does not affect variance, and scaling by 1/S_XX
scales variance by 1/S_XX squared. Conditional on the full X vector,
independent sampled pairs make Cov(u_i,u_j|X)=0 for i unequal j;
Var(u_i|X)=Var(u_i|X_i). The displayed result allows
heteroskedasticity because Var(u_i|X_i) need not be constant.
Zero conditional mean is needed for the regression model's OLS properties,
though the variance-of-a-sum algebra itself does not require it.
Assume S_XX>0 and finite relevant moments.
:::

## Replace the unknown error variation with residuals

We do not observe $u_i$ or $\operatorname{Var}(u_i\mid X_i)$. After fitting OLS, use
$\hat u_i=Y_i-\hat\beta_0-\hat\beta_1X_i$.

$$
\widehat{\operatorname{Var}}(\hat\beta_1)
=\underbrace{\frac{n}{n-2}}_{\text{degrees-of-freedom adjustment}}
\frac{\sum_i d_i^2\hat u_i^2}{S_{XX}^2},
\qquad
SE(\hat\beta_1)=\sqrt{\widehat{\operatorname{Var}}(\hat\beta_1)}.
$$

- The squared residuals stand in for unknown error variation.
- The $n/(n-2)$ factor adjusts for estimating two coefficients.
- This **heteroskedasticity-robust** standard error supports large-sample inference.

::: {.notes}
This is Equation 5.4, rearranged, often called HC1 in simple regression.
The residual plug-in and degrees-of-freedom correction do not make the
formula an exact unbiased conditional variance estimator in every finite
sample. Its justification is large-sample consistency under the maintained
conditions. R's default lm summary uses a homoskedastic-only standard error;
a robust covariance estimator is needed for this formula.
:::

## Three steps for a two-sided slope test

::: {.incremental}

1. Estimate $\hat\beta_1$ and its standard error.
2. Calculate $t=(\hat\beta_1-\beta_{1,0})/SE(\hat\beta_1)$.
3. Compare $|t|$ with the cutoff, or calculate the two-sided $p$-value.

:::

$$
\text{At }\alpha=0.05:\quad
|t|>1.96\ \Longrightarrow\ \text{reject }H_0.
$$

::: {.notes}
Key Concept 5.2. The cutoff uses a large-sample standard normal reference
distribution. Rejection is evidence under the maintained model assumptions.
:::

## What a two-sided $p$-value asks

For an observed statistic $t^{\mathrm{obs}}$, the large-sample $p$-value is

$$
p=\Pr\!\left(|Z|\ge |t^{\mathrm{obs}}|\right)
  =2\Phi\!\left(-|t^{\mathrm{obs}}|\right),
\qquad Z\sim N(0,1).
$$

- It is a **tail probability calculated under $H_0$**.
- Both tails count because the alternative allows positive or negative departures.
- Reject at 5% when $p<0.05$; equivalently, when $|t|>1.96$.

::: {.notes}
Equation 5.7. The p-value is not the probability that the null is true.
:::

## Read the California regression report

$$
\widehat{\text{TestScore}}=698.9-2.28\,\text{STR},
\qquad R^2=0.051,\quad SER=18.6.
$$

| Coefficient | Estimate | Standard error |
|:--|--:|--:|
| Intercept | $698.9$ | $10.4$ |
| STR slope | $-2.28$ | $0.52$ |

::: {.key-idea}
The slope's **$0.52$ standard error** is the number needed to test a claim about the slope.
:::

::: {.notes}
Stock and Watson, Equation 5.8. These are reported textbook values, rounded
for display. R-squared and SER measure fit, not slope uncertainty.
:::

## Test the claim that the slope is zero

$$
t^{\mathrm{obs}}=\frac{-2.28-0}{0.52}\approx-4.38.
$$

::: {.fragment}

$$|t^{\mathrm{obs}}|=4.38>1.96.$$

The null hypothesis $\beta_1=0$ is rejected at the **5% two-sided** level.

:::

::: {.notes}
The ratio uses rounded estimates. The test concerns the population regression
slope. Causal interpretation needs additional assumptions from Chapter 4.
:::

## See the two-sided rejection regions

```{r}
#| label: two-sided-tail-plot
#| echo: false
#| fig-width: 11
#| fig-height: 5
#| out-width: "85%"

test_statistic <- -2.28 / 0.52
normal_curve <- data.frame(z = seq(-5, 5, length.out = 1001))
normal_curve$density <- dnorm(normal_curve$z)
normal_curve$tail <- abs(normal_curve$z) >= 1.96

ggplot(normal_curve, aes(z, density)) +
  geom_area(
    data = subset(normal_curve, tail),
    aes(group = z < 0),
    fill = "#E85D4A",
    alpha = 0.8
  ) +
  geom_line(colour = "#242424", linewidth = 1.2) +
  geom_vline(
    xintercept = c(-1.96, 1.96),
    linetype = "dashed",
    colour = "#242424"
  ) +
  geom_vline(xintercept = test_statistic, colour = "#B83D2F", linewidth = 1.2) +
  annotate("text", x = -3.5, y = 0.11, label = "Observed t ≈ −4.38",
           colour = "#B83D2F", size = 5.5) +
  annotate("text", x = 0, y = 0.16, label = "95% non-rejection region",
           colour = "#242424", size = 5) +
  scale_x_continuous(breaks = c(-4.38, -1.96, 0, 1.96, 4.38)) +
  labs(x = "Standard normal statistic under the null", y = "Density") +
  theme(axis.title.y = element_blank())
```

::: {.notes}
The shaded areas begin at the two-sided 5% critical values. The solid red
line marks the observed statistic; its p-value corresponds to more extreme
areas beyond plus/minus 4.38. The curve is the approximate
reference distribution under the null.
:::

## A one-sided alternative changes the tail

If theory specifies a negative slope **before seeing the estimate**:

$$
H_0:\beta_1=0
\qquad\text{versus}\qquad
H_1:\beta_1<0.
$$

- The $t$-statistic remains $-4.38$.
- For a left-sided test, $p=\Phi(t^{\mathrm{obs}})$.
- At the 5% level, reject when $t<-1.645$.

::: {.definition}
A positive estimate, however large, does **not** support a negative-slope alternative.
:::

::: {.notes}
Equations 5.9–5.10. A right-sided alternative uses the right tail. Choose the
direction from a substantive prior reason, not after seeing the estimate.
:::

## Choose the test for the question

| Alternative | Evidence against $H_0$ | Large-sample 5% cutoff |
|:--|:--|:--|
| $\beta_1\ne\beta_{1,0}$ | Either direction | $|t|>1.96$ |
| $\beta_1<\beta_{1,0}$ | Negative direction | $t<-1.645$ |
| $\beta_1>\beta_{1,0}$ | Positive direction | $t>1.645$ |

::: {.exercise}
Suppose $t=+2.3$. Which alternatives reject at 5%?
:::

::: {.answer .fragment}
The **two-sided** and **right-sided** tests reject. The left-sided test does not.
:::

::: {.notes}
Check that students apply the direction as well as the size of a statistic.
:::

## The intercept follows the same procedure

For a proposed intercept $\beta_{0,0}$:

$$
H_0:\beta_0=\beta_{0,0},
\qquad
t=\frac{\hat\beta_0-\beta_{0,0}}{SE(\hat\beta_0)}.
$$

- Use the **intercept's** standard error; the California report gives $10.4$.
- Choose a two-sided or one-sided alternative according to the question.
- Apply the same large-sample $p$-value or critical-value rule.

::: {.notes}
Equation 5.11. Here the intercept is the fitted score at STR=0, well outside
the observed range, so many intercept hypotheses have little practical meaning.
:::


## Confidence intervals for regression coefficients {.section-slide background-color="#E85D4A"}

Section 5.2 · Which coefficient values remain consistent with the data?

::: {.notes}
Source: Stock and Watson, Chapter 5, Section 5.2, supplied in
book/chapter5. This section uses the same large-sample approximation and
standard errors as the hypothesis tests in Section 5.1.
:::

## A confidence interval collects plausible slope values

For each proposed value $b$, imagine testing $H_0:\beta_1=b$ against a two-sided alternative.

::: {.key-idea}
A **95% confidence interval** contains the values of $b$ that a 5% two-sided test would **not reject**.
:::

- Values close to $\hat\beta_1$ produce small absolute $t$-statistics.
- Values far from $\hat\beta_1$ produce large absolute $t$-statistics.
- The boundary occurs where $|t|=1.96$ in the large-sample approximation.

::: {.notes}
This is test inversion: one confidence interval summarizes the results of
testing every possible null value b. "Not rejected" does not mean the null
is proven true. The equivalence uses the same standard error and normal
critical value as the Section 5.1 test.
:::

## Solve the test rule for the unknown slope

A proposed value $b$ is retained at the 5% level when

$$
\left|\frac{\hat\beta_1-b}{SE(\hat\beta_1)}\right|\leq1.96.
$$

Multiply by the positive standard error and rearrange:

$$
-1.96\,SE(\hat\beta_1)
\leq\hat\beta_1-b
\leq1.96\,SE(\hat\beta_1)
$$

$$
\Longleftrightarrow\quad
\hat\beta_1-1.96\,SE(\hat\beta_1)
\leq b\leq
\hat\beta_1+1.96\,SE(\hat\beta_1).
$$

::: {.notes}
This derives Equation 5.12 from the non-rejection rule. The standard error
is positive and, for this simple test, does not depend on the proposed null
value b. The bounds are approximate because the 1.96 reference value comes
from the large-sample normal approximation.
:::

## The 95% interval is estimate ± uncertainty

$$
CI_{95\%}(\beta_1)=
\left[
\hat\beta_1-1.96\,SE(\hat\beta_1),\
\hat\beta_1+1.96\,SE(\hat\beta_1)
\right].
$$

::: columns
::: {.column width="48%"}

### Center

$\hat\beta_1$ is the estimated slope.

:::
::: {.column width="48%"}

### Margin

$1.96\,SE(\hat\beta_1)$ reflects sampling uncertainty.

:::
:::

::: {.key-idea}
A larger standard error makes the interval wider; a more precise estimate makes it narrower.
:::

::: {.notes}
Key Concept 5.3. The heteroskedasticity-robust SE from the prior section
can be used here. Do not confuse this interval for a population coefficient
with the SER or an interval for an individual future outcome.
:::

## What does 95% confidence mean?

Imagine drawing many new samples, fitting the same regression, and constructing a new interval each time.

- The **population slope $\beta_1$ stays fixed**.
- The estimate, standard error, and interval vary across samples.
- Under the maintained assumptions, **about 95% of these intervals cover $\beta_1$** in large samples.

::: {.definition}
After observing one interval, its endpoints are fixed. The 95% describes the **coverage of the method across repeated samples**.
:::

::: {.notes}
The textbook states the repeated-sampling interpretation. Avoid assigning a
95% probability to the fixed parameter conditional on this realized interval.
Coverage is approximate here because inference uses a large-sample reference
distribution and an estimated standard error. Biased sampling, a misspecified
model, or an invalid SE can spoil coverage.
:::

## Calculate the California slope interval

For the student–teacher ratio regression,
$\hat\beta_1=-2.28$ and $SE(\hat\beta_1)=0.52$.

$$
\begin{aligned}
\text{Margin}&=1.96(0.52)=1.0192,\\
CI_{95\%}(\beta_1)
&=[-2.28-1.0192,\,-2.28+1.0192]\\
&\approx\boxed{[-3.30,\,-1.26]}.
\end{aligned}
$$

- Both endpoints are in **test-score points per additional student per teacher**.
- Zero is outside the interval, agreeing with the 5% two-sided test.

::: {.notes}
Stock and Watson's California example following Equation 5.12. The
coefficient and SE are rounded textbook figures, so the displayed bounds
are approximate. The negative association is statistically distinguishable
from zero under the maintained assumptions; the interval alone does not
establish a causal class-size effect.
:::

## The intercept can have an interval too

Replace the slope estimate and standard error with those for the intercept:

$$
CI_{95\%}(\beta_0)=
\left[
\hat\beta_0-1.96\,SE(\hat\beta_0),\
\hat\beta_0+1.96\,SE(\hat\beta_0)
\right].
$$

For the California regression, $698.9\pm1.96(10.4)$ gives approximately **$[678.5,\,719.3]$**.

::: {.key-idea}
The calculation is valid under the model, but the intercept is the predicted score at $\text{STR}=0$, far outside a meaningful district comparison.
:::

::: {.notes}
Section 5.2 says the same construction applies to beta_0. The numerical
intercept interval uses the reported estimates in Equation 5.8 and rounds
the endpoints to one decimal place. Avoid treating extrapolation to STR=0
as a useful policy estimate.
:::

## Translate a slope interval into a change in X

In the linear conditional-mean model, a change of $\Delta x$ gives

$$
E(Y\mid X=x+\Delta x)-E(Y\mid X=x)=\beta_1\Delta x.
$$

If $CI_{95\%}(\beta_1)=[L,U]$, multiply **both endpoints** by $\Delta x$:

$$
CI_{95\%}(\beta_1\Delta x)
=
[\min(L\Delta x,U\Delta x),\,
 \max(L\Delta x,U\Delta x)].
$$

::: {.definition}
When $\Delta x<0$, multiplication reverses the order of the endpoints.
:::

::: {.notes}
This is Equation 5.13 with the interval reordered when the change is
negative. The mapping has the same approximate coverage because multiplying
by a known nonzero constant is one-to-one. This is an interval for a model
mean difference, not an individual prediction interval. A causal
interpretation requires the identifying assumptions discussed in Lecture 2.
:::

## Two fewer students per teacher

Suppose the student–teacher ratio falls by two: $\Delta x=-2$.

$$
\widehat{\Delta\text{TestScore}}
=(-2.28)(-2)=4.56\text{ points}.
$$

The slope interval $[-3.30,-1.26]$ becomes

$$
[(-1.26)(-2),\,(-3.30)(-2)]
=\boxed{[2.52,\,6.60]\text{ points}}.
$$

::: {.key-idea}
The model predicts a mean score increase between **2.52 and 6.60 points** at the 95% confidence level.
:::

::: {.notes}
Stock and Watson's final example in Section 5.2. The endpoints use the
rounded slope interval; applying -2 to the unrounded bounds leads to the
same values at two decimals. The interval describes uncertainty about the
model's mean difference. It does not describe the range of scores for
individual districts and is causal only if the slope has a credible causal
interpretation.
:::
