---
title: "Lecture 1: Asking Questions with Data"
subtitle: "Economic questions, data, and the probability toolkit"
author: "Swapnil Singh"
date: "2026-09-01"
date-format: "D MMMM YYYY"
footer: "Topics in Econometrics I · Lecture 1"
knitr:
  opts_chunk:
    out.width: "80%"
---

```{r}
#| label: setup
#| include: false

library(ggplot2)

set.seed(2026)

course_red <- "#E85D4A"
course_red_dark <- "#B83D2F"
course_charcoal <- "#242424"
course_grey <- "#DEDBD6"
course_pale <- "#FBE9E5"

theme_set(
  theme_minimal(base_size = 18) +
    theme(
      plot.title = element_text(
        colour = course_charcoal,
        face = "bold"
      ),
      plot.subtitle = element_text(colour = "#68645F"),
      panel.grid.minor = element_blank(),
      strip.text = element_text(face = "bold"),
      plot.background = element_rect(fill = "transparent", colour = NA)
    )
)

# Simulated school-district data: the data-generating process is known here.
n_districts <- 240
school_data <- data.frame(
  district = seq_len(n_districts),
  advantage = rnorm(n_districts)
)
school_data$class_size <- 23 - 2.2 * school_data$advantage +
  rnorm(n_districts, sd = 1.8)
school_data$test_score <- 665 - 0.7 * school_data$class_size +
  11 * school_data$advantage + rnorm(n_districts, sd = 5.5)
school_data$advantage_group <- cut(
  school_data$advantage,
  breaks = quantile(school_data$advantage, probs = c(0, 1 / 3, 2 / 3, 1)),
  include.lowest = TRUE,
  labels = c("Lower", "Middle", "Higher")
)

# A small illustrative panel.
panel_data <- expand.grid(
  year = 2019:2023,
  city = c("Aster", "Birch", "Cedar", "Dover", "Elm", "Flint")
)
city_effect <- c(Aster = 6, Birch = 2, Cedar = -2, Dover = 4, Elm = -5, Flint = 0)
panel_data$ridership <- 54 + city_effect[panel_data$city] +
  2.2 * (panel_data$year - 2019) + rnorm(nrow(panel_data), sd = 1.8)

# Objects used for probability illustrations.
failure_distribution <- data.frame(
  failures = 0:4,
  probability = c(0.80, 0.10, 0.06, 0.03, 0.01)
)

commute_joint <- data.frame(
  Weather = c("Rain", "No rain", "Total"),
  `Long commute` = c(0.15, 0.07, 0.22),
  `Short commute` = c(0.15, 0.63, 0.78),
  Total = c(0.30, 0.70, 1.00),
  check.names = FALSE
)
```

## Before we begin

::: {.key-idea}
**Econometrics is disciplined learning from imperfect data.**

We will keep asking: *What can the data tell us—and what must come from assumptions?*
:::

::: {.incremental}

- Bring a question, not merely a dataset.
- Translate words into a measurable target.
- Treat uncertainty as part of the answer.

:::

::: {.notes}
Timing: 2 minutes. Ask for one example of a quantitative economic question from the room. Keep the answer for the final slide.
:::

## By the end, you should be able to…

::: {.incremental}

1. distinguish **description**, **prediction**, and **causal inference**;
2. recognize cross-sectional, time-series, and panel data;
3. interpret a probability distribution, mean, variance, and conditional mean;
4. explain sampling variation, the law of large numbers, and the central limit theorem;
5. say what extra assumptions a claim needs.

:::

::: {.notes}
Timing: 2 minutes. Return to these outcomes in the exit ticket.
:::

## Part I · Economic questions {.section-slide background-color="#E85D4A"}

Good analysis begins before the first line of code.

::: {.notes}
Clock: 00:00–00:35, including the opening and a four-minute design activity.
:::

## Some questions

::: columns
::: {.column width="48%"}

### Education

How much would smaller classes improve achievement?

### Lending

Are otherwise similar applicants treated differently by race?

:::
::: {.column width="48%"}

### Public health

How much would a cigarette tax reduce smoking?

### Macroeconomy

How fast will GDP grow next year?

:::
:::

::: {.notes}
Timing: 2 minutes. These are the four motivating questions in Chapter 1. Ask which one is different in its inferential goal: GDP forecasting is predictive; the first three are causal.
:::

## Same data, different goals

::: columns
::: {.column width="31%"}

### Describe

What is the average test score among districts in our data?

**Summarize what we observed.**

:::
::: {.column width="31%"}

### Predict

For a new district with 22 students per class, what score should we expect?

**Use known characteristics to forecast an outcome.**

:::
::: {.column width="31%"}

### Explain a cause

If a district reduced class size, how much would its score change?

**Compare what would happen with and without the change.**

:::
:::

::: {.key-idea .fragment}
A model does not choose the goal. **The question does.**
:::

::: {.notes}
Timing: 3 minutes. Use the same class-size setting for all three goals. No notation is needed yet; potential outcomes arrive on the next causal slide.
:::

## The chain of an econometric answer

::: {.incremental}

**Question** → What decision or comparison matters?

**Estimand** → What population quantity would answer it?

**Data** → What was measured, on whom, and when?

**Assumptions** → Why does the calculation target the estimand?

**Estimate + uncertainty** → What did this sample teach us?

:::

::: {.notes}
Timing: 2 minutes. Say “estimand” slowly: the target, not the estimate. Use the distinction between a population average and a sample average.
:::

## A tempting pattern

```{r}
#| label: raw-class-size
#| echo: false
#| fig-width: 10
#| fig-height: 5.7
#| out-width: "78%"

ggplot(school_data, aes(class_size, test_score)) +
  geom_point(colour = course_red, alpha = 0.55, size = 2.4) +
  geom_smooth(
    method = "lm", se = FALSE,
    colour = course_charcoal, linewidth = 1.3
  ) +
  labs(
    title = "Districts with smaller classes have higher scores",
    subtitle = "Simulated cross-sectional data",
    x = "Students per class",
    y = "Average test score"
  )
```

::: {.notes}
Timing: 2 minutes. Ask: “Does this line tell us what would happen if a district hired teachers tomorrow?” Do not reveal the simulated data-generating process yet.
:::

## What comparison do we want?

For district $i$, imagine two outcomes:

$$
Y_i(1)=\text{score with a smaller class},\qquad
Y_i(0)=\text{score without it}.
$$

::: {.definition .fragment}
The individual causal effect is $Y_i(1)-Y_i(0)$. The average causal effect is

$$\operatorname{ATE}=E[Y_i(1)-Y_i(0)].$$
:::

::: {.key-idea .fragment}
**The missing counterfactual:** for the same district at the same moment, we observe only one of these two outcomes—never both.
:::

::: {.notes}
Timing: 3 minutes. Call this the missing counterfactual, not a missing-data problem that software can simply fill in.
:::

## Why random assignment is powerful

::: columns
::: {.column width="46%"}

### Treatment group

Randomly assigned to smaller classes

::: {.comparison-value}
**Average test score**
:::

::: {.comparison-caption}
smaller classes
:::

:::
::: {.column width="8%"}

###

**vs.**

:::
::: {.column width="46%"}

### Control group

Randomly assigned to usual classes

::: {.comparison-value}
**Average test score**
:::

::: {.comparison-caption}
usual classes
:::

:::
:::

::: {.key-idea .fragment}
With successful random assignment, the groups are comparable *before* treatment. The difference between their average scores estimates the causal effect.
:::

::: {.notes}
Timing: 2 minutes. Stress “in expectation” and “with a sufficiently large sample.” Randomization balances both observed and unobserved pre-treatment characteristics on average.
:::

## The ideal experiment is a benchmark

::: {.incremental}

- **Ethics:** We cannot randomly assign people to smoke.
- **Feasibility:** A multi-year school experiment is expensive and difficult.
- **Compliance:** Assigned treatment may differ from received treatment.
- **External validity:** Volunteers or study sites may not represent the target population.

:::

::: {.key-idea .fragment}
When an experiment is unavailable, ask: **What would the ideal experiment have compared?**
:::

::: {.notes}
Timing: 2 minutes. Connect to the Tennessee class-size experiment mentioned in the source chapter, without dwelling on historical details.
:::

## The third variable: district advantage

```{r}
#| label: confounding-plot
#| echo: false
#| fig-width: 10
#| fig-height: 5.7
#| out-width: "78%"

ggplot(school_data, aes(class_size, test_score, colour = advantage_group)) +
  geom_point(alpha = 0.50, size = 2.2) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.05) +
  scale_colour_manual(
    values = c("Lower" = course_red_dark, "Middle" = "#8B8580", "Higher" = course_charcoal),
    labels = c(
      "Lower" = "Lower advantage",
      "Middle" = "Middle advantage",
      "Higher" = "Higher advantage"
    ),
    name = "District advantage"
  ) +
  labs(
    title = "District advantage affects both variables",
    subtitle = "Higher family resources → smaller classes and higher scores (simulated)",
    x = "Students per class",
    y = "Average test score"
  )
```

::: {.notes}
Timing: 2 minutes. Because these data are simulated, we know the mechanism: district advantage affects class sizes and outcomes. In real data, the mechanism must be argued and investigated.
:::

## Association is not automatically causation

::: columns
::: {.column width="44%"}

$$
\text{District advantage}
\longrightarrow \text{Class size}
$$

$$
\text{District advantage}
\longrightarrow \text{Test score}
$$

:::
::: {.column width="52%"}

::: {.definition}
A **confounder** helps determine both the treatment and the outcome, creating a comparison between unlike units.
:::

:::
:::

::: {.key-idea .fragment}
Regression can adjust for measured differences. It cannot guarantee that every important difference was measured correctly.
:::

::: {.notes}
Timing: 2 minutes. This is a verbal causal diagram. Avoid claiming that “controlling for everything” solves identification.
:::

## Prediction does not require causation

::: {.exercise}
You see many umbrellas outside the library. Is “umbrella use” useful for predicting rain? Does opening an umbrella cause rain?
:::

::: {.answer .fragment}
Umbrellas can be highly predictive without being a cause of rain. Prediction needs a stable relationship; causal inference needs a credible comparison under intervention.
:::

::: {.notes}
Timing: 1 minute. This is the chapter’s core umbrella intuition. Elicit the answer before revealing it.
:::

## Active check · Design the comparison

::: {.exercise}
**Pairs · 4 minutes**

Question: *Does one additional hour of studying improve the next econometrics test score?*

1. State the treatment and outcome.
2. Describe an ideal randomized experiment.
3. Name one practical or ethical problem.
4. What would a purely predictive version of the question be?
:::

::: {.notes}
Timing: 4 minutes for pairs. Listen for treatment versions: “one hour” must be defined. Note spillovers and noncompliance if students raise them.
:::

## Design check · Debrief

::: {.answer}

- **Treatment:** a defined additional hour of study support before the test.
- **Outcome:** score on a specified test.
- **Ideal comparison:** randomly offer the support, then compare group means.
- **Complications:** compliance, spillovers, heterogeneous study quality, fairness.
- **Prediction:** “Given prior grades and study hours, what score should we expect?”

:::

::: {.key-idea .fragment}
The causal and predictive questions may use the same variables—and still require different logic.
:::

::: {.notes}
Timing: 1 minute for the debrief.
:::

## Part II · What kind of data? {.section-slide background-color="#B83D2F"}

Rows and columns have histories.

::: {.notes}
Clock: 00:35–00:55, including a short classification check.
:::

## Where did the data come from?

::: columns
::: {.column width="48%"}

### Experimental

Treatment is assigned by a designed mechanism.

**Strength:** comparability can come from randomization.

:::
::: {.column width="48%"}

### Observational

Behavior is recorded as it occurs through surveys, transactions, or administrative systems.

**Challenge:** treatment choices are usually not random.

:::
:::

::: {.small-note .fragment}
“Observational” does not mean “bad”; it means the design burden is different.
:::

::: {.notes}
Timing: 2 minutes. Name measurement error and selection as issues in either source; randomization does not fix a poorly measured outcome.
:::

## The shape of a dataset

| Data structure | Entities | Times | Main variation |
|---|---:|---:|---|
| Cross-section | many | one | across entities |
| Time series | one | many | across time |
| Panel / longitudinal | many | many | both dimensions |

::: {.key-idea .fragment}
The data structure determines which comparisons are available—and which uncertainty calculations are credible.
:::

::: {.notes}
Timing: 2 minutes. Introduce $n$ for entities and $T$ for periods. Mention repeated cross-sections as a fourth common structure, but do not develop it here.
:::

## Cross-section · many entities, one period

```{r}
#| label: cross-section-table
#| echo: false

school_snapshot <- school_data[c(7, 31, 88, 143, 219),
  c("district", "test_score", "class_size", "advantage_group")]
school_snapshot$test_score <- round(school_snapshot$test_score, 1)
school_snapshot$class_size <- round(school_snapshot$class_size, 1)
names(school_snapshot) <- c("District", "Test score", "Class size", "Advantage")

knitr::kable(school_snapshot, align = c("r", "r", "r", "l"))
```

::: {.definition}
Each row is a different district measured in the same period. Row order has no economic meaning.
:::

::: {.notes}
Timing: 2 minutes. Relate to the California district data in Chapter 1: $n=420$ districts in 1999.
:::

## Time series · one entity, many periods

```{r}
#| label: time-series-plot
#| echo: false
#| fig-width: 10
#| fig-height: 5.5

air_data <- data.frame(
  month = as.Date(paste0(time(AirPassengers), "-01"), format = "%Y.%m-%d"),
  passengers = as.numeric(AirPassengers)
)

ggplot(air_data, aes(month, passengers)) +
  geom_line(colour = course_charcoal, linewidth = 1) +
  geom_point(colour = course_red, size = 1.1, alpha = 0.65) +
  labs(
    title = "One system observed repeatedly",
    subtitle = "International airline passengers, 1949–1960 (built-in R series)",
    x = NULL,
    y = "Passengers (thousands)"
  )
```

::: {.notes}
Timing: 2 minutes. Point out trend, seasonality, and dependence across adjacent months. Time order matters, unlike row order in a cross-section.
:::

## Panel · many entities, many periods

```{r}
#| label: panel-plot
#| echo: false
#| fig-width: 10
#| fig-height: 5.4

ggplot(panel_data, aes(year, ridership, group = city, colour = city)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2.2) +
  scale_x_continuous(breaks = 2019:2023) +
  scale_colour_manual(
    values = c(course_red, course_charcoal, "#817C77", "#BC766B", "#4E4B48", "#D9A199")
  ) +
  labs(
    title = "Compare cities—and compare each city with itself",
    subtitle = "Illustrative public-transport ridership index",
    x = NULL,
    y = "Ridership index",
    colour = "City"
  )
```

::: {.notes}
Timing: 2 minutes. Relate to the chapter’s state-by-year cigarette data: $n=48$, $T=11$, and 528 state-year observations.
:::

## Active check · Name the structure

::: {.exercise}
Classify each dataset.

1. Wages of 2,000 workers surveyed in March 2026.
2. Lithuania’s quarterly GDP from 2000 to 2025.
3. Annual emissions for 27 EU countries from 1995 to 2025.
4. A new sample of households surveyed every year from 2010 to 2025.
:::

::: {.answer .fragment}
1. Cross-section · 2. Time series · 3. Panel · 4. Repeated cross-sections, unless the same households are followed.
:::

::: {.notes}
Timing: 3 minutes. The fourth item checks whether students track entities, not merely repeated dates.
:::

## Pause · 15 minutes {.section-slide background-color="#E85D4A"}

When we return: turning randomness into quantities.

::: {.notes}
Clock: 00:55–01:10. Leave this slide visible during the pause.
:::

## Part III · Probability distributions {.section-slide background-color="#E85D4A"}

First describe the possible values and how probability is distributed across them.

::: {.notes}
Clock: 01:10–01:35.
:::

## Why probability appears in econometrics

We observe one sample, but a different sample would give a different answer.

::: {.incremental}

- Which outcomes were possible?
- How frequently would each occur in repeated sampling?
- How much does an estimate vary across samples?
- Is the observed estimate surprising under a proposed model?

:::

::: {.key-idea .fragment}
Probability connects the dataset we have to the samples we might have drawn.
:::

::: {.notes}
Timing: 2 minutes. Distinguish empirical uncertainty from vagueness. The model describes a repeatable sampling thought experiment.
:::

## Outcome → random variable

::: columns
::: {.column width="48%"}

### Random process

A student commutes to campus.

Possible outcomes include 14.2, 19.7, or 31.4 minutes.

:::
::: {.column width="48%"}

### Numerical summary

Let $Y$ be commute time in minutes.

$Y$ is a **random variable** because its realized value is not known beforehand.

:::
:::

::: {.key-idea .fragment}
**Two types:** discrete random variables have countable values; continuous random variables can take any value in an interval.
:::

::: {.notes}
Timing: 2 minutes. Emphasize that the random variable is a mapping from outcomes to numbers; it is not “a variable with errors.”
:::

## A discrete probability distribution

```{r}
#| label: failure-distribution-plot
#| echo: false
#| fig-width: 9.5
#| fig-height: 5.6
#| out-width: "75%"

ggplot(failure_distribution, aes(failures, probability)) +
  geom_col(fill = course_red, width = 0.68) +
  geom_text(
    aes(label = scales::percent(probability, accuracy = 1)),
    vjust = -0.35,
    colour = course_charcoal,
    size = 6
  ) +
  scale_x_continuous(breaks = 0:4) +
  scale_y_continuous(
    limits = c(0, 0.90),
    labels = scales::percent_format()
  ) +
  labs(
    title = "Possible network failures and their probabilities",
    x = "Number of failures, M",
    y = "Probability"
  )
```

::: {.notes}
Timing: 2 minutes. Check that probabilities are nonnegative and sum to one. This adapts the wireless-network example in Chapter 2.
:::

## Bernoulli distribution · yes or no

Let $D$ record whether an event occurs:

$$
D\in\{0,1\},\qquad
\Pr(D=1)=p,\qquad
\Pr(D=0)=1-p.
$$

::: columns
::: {.column width="48%"}

### $D=1$

The event occurs: employed, loan denied, treatment received.

:::
::: {.column width="48%"}

### $D=0$

The event does not occur: not employed, approved, untreated.

:::
:::

::: {.key-idea .fragment}
One number, $p$, completely describes a Bernoulli distribution.
:::

::: {.notes}
Timing: 2 minutes. Keep this slide about the distribution only. Its mean and variance are derived after those concepts have been defined.
:::

## Continuous probability distributions

A continuous variable can take any value in an interval. Its distribution is described by a **density**, $f_Y(y)$.

$$
\Pr(a\le Y\le b)=\int_a^b f_Y(y)\,dy.
$$

::: columns
::: {.column width="48%"}

### Probability is area

The total area under the density is one.

:::
::: {.column width="48%"}

### A point has no area

For a continuous variable, $\Pr(Y=y)=0$ at any exact value.

:::
:::

::: {.notes}
Timing: 2 minutes. Contrast a density height with a probability. The next two slides show two continuous distributions with very different shapes.
:::

## Normal distribution · symmetric and continuous

```{r}
#| label: normal-distribution-plot
#| echo: false
#| fig-width: 10
#| fig-height: 5.2
#| out-width: "74%"

normal_grid <- data.frame(score = seq(35, 105, length.out = 600))
normal_grid$density <- dnorm(normal_grid$score, mean = 70, sd = 10)
normal_grid$inside <- normal_grid$score >= 70 - 1.96 * 10 &
  normal_grid$score <= 70 + 1.96 * 10

ggplot(normal_grid, aes(score, density)) +
  geom_area(
    data = subset(normal_grid, inside),
    fill = course_red,
    alpha = 0.45
  ) +
  geom_line(colour = course_charcoal, linewidth = 1.2) +
  geom_vline(xintercept = 70, linetype = "dashed", colour = course_red_dark) +
  labs(
    title = "A bell shape centered at the mean",
    subtitle = "For N(70, 10²), 95% of probability lies between 50.4 and 89.6",
    x = "Test score",
    y = "Density"
  )
```

::: {.notes}
Timing: 3 minutes. Introduce the notation $N(\mu,\sigma^2)$. The normal distribution is symmetric, has skewness zero, and kurtosis three.
:::

## Log-normal distribution · positive and right-skewed

```{r}
#| label: lognormal-distribution-plot
#| echo: false
#| fig-width: 10
#| fig-height: 5.2
#| out-width: "74%"

lognormal_grid <- data.frame(
  minutes = seq(5, 50, length.out = 600)
)
lognormal_grid$density <- dlnorm(
  lognormal_grid$minutes,
  meanlog = log(20),
  sdlog = 0.35
)
lognormal_grid$inside <- lognormal_grid$minutes >= 15 &
  lognormal_grid$minutes <= 25

ggplot(lognormal_grid, aes(minutes, density)) +
  geom_area(
    data = subset(lognormal_grid, inside),
    fill = course_red,
    alpha = 0.50
  ) +
  geom_line(colour = course_charcoal, linewidth = 1.2) +
  labs(
    title = "A long right tail represents occasional large values",
    subtitle = "Shaded: probability that commute time is between 15 and 25 minutes",
    x = "Commute time (minutes)",
    y = "Density"
  )
```

::: {.notes}
Timing: 3 minutes. A log-normal variable is positive because its logarithm is normal. Earnings, prices, firm size, and durations are often right-skewed, although log-normality still must be checked.
:::

## Distribution families we will meet

| Distribution | Values | Shape and later role |
|---|---|---|
| Bernoulli | $0$ or $1$ | Binary outcomes and treatment indicators |
| Normal | $(-\infty,\infty)$ | Symmetric benchmark; approximates many averages |
| Log-normal | $(0,\infty)$ | Right-skewed model for positive quantities |
| Chi-squared, $\chi_m^2$ | $[0,\infty)$ | Sums of squared standard normals |
| Student $t_m$ | $(-\infty,\infty)$ | Bell-shaped with heavier tails than normal |
| $F_{m,n}$ | $[0,\infty)$ | Ratios of scaled chi-squared variables |

::: {.small-note}
The degrees of freedom $m$ and $n$ control the shapes of $\chi^2$, $t$, and $F$ distributions.
:::

::: {.notes}
Timing: 3 minutes. The last three distributions support hypothesis tests later in the course. For now, students should recognize their support, shape, and names.
:::

## Part IV · Moments and relationships {.section-slide background-color="#B83D2F"}

Now summarize a distribution’s center, spread, shape, and co-movement.

::: {.notes}
Clock: 01:35–02:00.
:::

## Mean · the center of a distribution

The **mean**, **expected value**, and **expectation** are the same population quantity, denoted $\mu_Y=E(Y)$.

::: columns
::: {.column width="48%"}

### Discrete $Y$

$$
E(Y)=\sum_y y\Pr(Y=y)
$$

:::
::: {.column width="48%"}

### Continuous $Y$

$$
E(Y)=\int_{-\infty}^{\infty} y f_Y(y)\,dy
$$

:::
:::

::: {.definition .fragment}
The mean is the **probability-weighted average** of all possible values—the long-run average across repeated draws.
:::

::: {.notes}
Timing: 3 minutes. This is the first moment. Stress that a mean need not be a possible realization and that both formulas express the same idea.
:::

## Mean · a worked discrete example

For network failures $M$,

$$
E(M)=\sum_m m\Pr(M=m)
$$

$$
=0(0.80)+1(0.10)+2(0.06)+3(0.03)+4(0.01)=\mathbf{0.35}.
$$

::: {.key-idea .fragment}
$E(M)=0.35$ does **not** predict 0.35 failures in one session. It is the average over many sessions.
:::

::: {.notes}
Timing: 2 minutes. Ask whether 0.35 must be a possible realization. For a Bernoulli variable, use the same formula to derive $E(D)=p$.
:::

## Variance · how far outcomes wander

$$
\operatorname{Var}(Y)=E[(Y-\mu_Y)^2],
\qquad
\operatorname{SD}(Y)=\sqrt{\operatorname{Var}(Y)}.
$$

::: columns
::: {.column width="48%"}

### Variance

Squares deviations, so large misses matter more.

Units are squared.

:::
::: {.column width="48%"}

### Standard deviation

Returns to the original units of $Y$.

Think “typical scale of variation,” not a guaranteed distance.

:::
:::

::: {.key-idea .fragment}
For a Bernoulli variable, $E(D)=p$ and $\operatorname{Var}(D)=p(1-p)$.
:::

::: {.notes}
Timing: 2 minutes. For the network example, variance is about 0.648 and SD about 0.80 failures.
:::

## What happens when units change?

If $Z=a+bY$, then

$$
E(Z)=a+bE(Y),
$$

$$
\operatorname{Var}(Z)=b^2\operatorname{Var}(Y),
\qquad
\operatorname{SD}(Z)=|b|\operatorname{SD}(Y).
$$

::: {.exercise .fragment}
If temperature in Celsius has mean $20$ and SD $4$, what are the mean and SD in Fahrenheit, where $F=32+1.8C$?
:::

::: {.answer .fragment}
$E(F)=68$ and $\operatorname{SD}(F)=7.2$. Adding 32 moves the center but does not change the spread.
:::

::: {.notes}
Timing: 3 minutes. Pause for mental arithmetic before revealing. Stress the absolute value for a negative scale factor.
:::

## Skewness · asymmetry

$$
\operatorname{Skew}(Y)=
\frac{E[(Y-\mu_Y)^3]}{\sigma_Y^3}.
$$

::: columns
::: {.column width="31%"}

### Negative

Longer left tail

:::
::: {.column width="31%"}

### Zero

Symmetric distribution

:::
::: {.column width="31%"}

### Positive

Longer right tail

:::
:::

::: {.key-idea .fragment}
Cubing preserves the sign of deviations; dividing by $\sigma_Y^3$ removes the units.
:::

::: {.notes}
Timing: 2 minutes. A symmetric distribution has skewness zero, but zero skewness alone does not guarantee full symmetry.
:::

## Kurtosis · tail weight

$$
\operatorname{Kurt}(Y)=
\frac{E[(Y-\mu_Y)^4]}{\sigma_Y^4}.
$$

::: {.incremental}

- Fourth powers give extreme deviations substantial weight.
- Larger kurtosis means more probability in the tails and greater outlier risk.
- The normal distribution has kurtosis $3$.
- A distribution with kurtosis above $3$ is **heavy-tailed** relative to the normal.

:::

::: {.notes}
Timing: 2 minutes. Use the book’s Wall Street example: a 21-standard-deviation event is essentially impossible under a constant-variance normal model, revealing much heavier empirical tails.
:::

## Same center and spread, different shapes

```{r}
#| label: distribution-shapes-plot
#| echo: false
#| fig-width: 11
#| fig-height: 4.9
#| out-width: "82%"

standardize_draws <- function(values) {
  as.numeric((values - mean(values)) / sd(values))
}

shape_data <- rbind(
  data.frame(
    value = standardize_draws(rnorm(12000)),
    family = "Normal\nsymmetric; kurtosis 3"
  ),
  data.frame(
    value = standardize_draws(rlnorm(12000, sdlog = 0.55)),
    family = "Log-normal\nright-skewed"
  ),
  data.frame(
    value = standardize_draws(rt(12000, df = 5)),
    family = "Student t₅\nsymmetric; heavy tails"
  )
)
shape_data$family <- factor(
  shape_data$family,
  levels = c(
    "Normal\nsymmetric; kurtosis 3",
    "Log-normal\nright-skewed",
    "Student t₅\nsymmetric; heavy tails"
  )
)

ggplot(shape_data, aes(value)) +
  geom_density(fill = course_pale, colour = course_red_dark, linewidth = 0.9) +
  facet_wrap(~family, nrow = 1) +
  coord_cartesian(xlim = c(-4, 6), ylim = c(0, 0.55)) +
  labs(
    title = "All three are standardized to mean 0 and standard deviation 1",
    x = "Standardized value",
    y = "Density"
  )
```

::: {.notes}
Timing: 3 minutes. Equal means and variances do not imply equal distributions. Connect the right-skewed panel to earnings and the heavy-tailed panel to financial returns.
:::

## Standardization gives a common scale

$$
Z=\frac{Y-\mu_Y}{\sigma_Y}
$$

::: columns
::: {.column width="48%"}

### Center

$E(Z)=0$

:::
::: {.column width="48%"}

### Scale

$\operatorname{Var}(Z)=1$

:::
:::

::: {.definition .fragment}
A $z$-score says how many standard deviations an observation lies above or below the mean.
:::

::: {.notes}
Timing: 2 minutes. Standardization changes units, not rank. It will reappear in the CLT and test statistics.
:::

## Moments · a compact map

| Quantity | Formula | What it describes |
|---|---|---|
| Mean | $E(Y)$ | Center |
| Variance | $E[(Y-\mu_Y)^2]$ | Spread |
| Skewness | $E[(Y-\mu_Y)^3]/\sigma_Y^3$ | Asymmetry |
| Kurtosis | $E[(Y-\mu_Y)^4]/\sigma_Y^4$ | Tail weight |

::: {.key-idea .fragment}
Mean and variance do not fully describe a distribution; shape and tail behavior can matter economically.
:::

::: {.notes}
Timing: 2 minutes. The raw $r$th moment is $E(Y^r)$. Variance, skewness, and kurtosis use centered or standardized moments.
:::

## From one variable to two {.section-slide background-color="#B83D2F"}

Joint and conditional distributions describe economic relationships.

::: {.notes}
Continue Part IV. The mean and variance are now defined, so conditional means and covariance have a clear foundation.
:::

## Joint and marginal distributions

```{r}
#| label: commute-joint-table
#| echo: false

commute_display <- commute_joint
for (column in 2:4) {
  commute_display[[column]] <- sprintf("%.2f", commute_display[[column]])
}
knitr::kable(commute_display, align = c("l", "r", "r", "r"))
```

::: {.incremental}

- Interior cells are **joint probabilities**.
- Row and column totals are **marginal probabilities**.
- All four joint probabilities sum to one.

:::

::: {.notes}
Timing: 2 minutes. This adapts the rainy-commute table in Chapter 2. Ask students to locate Pr(rain and long commute) and Pr(long commute).
:::

## Conditioning changes the reference group

What is the probability of a long commute **given rain**?

$$
\Pr(\text{long}\mid\text{rain})
=\frac{\Pr(\text{long and rain})}{\Pr(\text{rain})}
=\frac{0.15}{0.30}=0.50.
$$

::: {.definition .fragment}
Conditioning restricts attention to cases where the condition is true, then renormalizes probabilities within that group.
:::

::: {.notes}
Timing: 3 minutes. The denominator is the new reference population. Contrast Pr(long | rain) with Pr(rain | long).
:::

## Independence is a strong statement

$X$ and $Y$ are independent when learning $X$ does not change the distribution of $Y$:

$$
\Pr(Y=y\mid X=x)=\Pr(Y=y)
$$

or, equivalently,

$$
\Pr(X=x,Y=y)=\Pr(X=x)\Pr(Y=y).
$$

::: {.exercise .fragment}
Are rain and commute length independent in the table?
:::

::: {.answer .fragment}
No. $\Pr(\text{long}\mid\text{rain})=0.50$, while $\Pr(\text{long})=0.22$.
:::

::: {.notes}
Timing: 2 minutes. Independence concerns the full distribution, not merely a zero correlation.
:::

## Conditional mean · center within a group

$$
m(x)=E(Y\mid X=x)
$$

::: {.incremental}

- It is the mean of $Y$ among cases with $X=x$.
- It summarizes how the center of $Y$ changes with information $X$.
- It can vary with $x$, tracing a **conditional mean function**.

:::

::: {.key-idea .fragment}
Regression is, at its core, a way to model a conditional mean.
:::

::: {.notes}
Timing: 2 minutes. This prepares the language for later OLS. Mention in delivery, but do not prove, that the conditional mean minimizes mean squared prediction error.
:::

## The law of iterated expectations

Suppose an old network averages $0.56$ failures and a new one averages $0.14$. Each is used half the time.

$$
E(M)=E[E(M\mid A)]
$$

::: {.fragment}
$$
=0.56(0.50)+0.14(0.50)=0.35.
$$
:::

::: {.key-idea .fragment}
The overall mean is a weighted average of group-specific means.
:::

::: {.notes}
Timing: 2 minutes. Ask what changes if 80% of sessions use the new network. The conditional means stay fixed; the overall mean changes through the weights.
:::

## Covariance · unscaled co-movement

$$
\operatorname{Cov}(X,Y)=E[(X-\mu_X)(Y-\mu_Y)]
$$

::: columns
::: {.column width="48%"}

**Positive:** above-average $X$ tends to accompany above-average $Y$.

:::
::: {.column width="48%"}

**Negative:** above-average $X$ tends to accompany below-average $Y$.

:::
:::

::: {.definition .fragment}
Covariance is positive when $X$ and $Y$ tend to deviate from their means in the same direction. Its units are the units of $X$ multiplied by the units of $Y$.
:::

::: {.notes}
Timing: 2 minutes. Use the signs of the centered products for intuition.
:::

## Correlation · unit-free linear co-movement

$$
\operatorname{Corr}(X,Y)=
\frac{\operatorname{Cov}(X,Y)}{\sigma_X\sigma_Y},
\qquad -1\le\operatorname{Corr}(X,Y)\le1.
$$

::: {.incremental}

- Dividing by both standard deviations removes the units.
- The sign gives the direction of **linear** co-movement.
- The magnitude describes the strength of that linear pattern.
- Correlation does not by itself identify a causal effect.

:::

::: {.notes}
Timing: 2 minutes. Perfect correlation of plus or minus one means all probability lies on a line with positive or negative slope; zero correlation need not mean independence.
:::

## Zero correlation can hide a relationship

```{r}
#| label: nonlinear-correlation
#| echo: false
#| fig-width: 9.8
#| fig-height: 5.4

x_nonlinear <- runif(500, -2, 2)
nonlinear_data <- data.frame(
  x = x_nonlinear,
  y = x_nonlinear^2 + rnorm(500, sd = 0.35)
)
nonlinear_correlation <- cor(nonlinear_data$x, nonlinear_data$y)

ggplot(nonlinear_data, aes(x, y)) +
  geom_point(colour = course_red, alpha = 0.45, size = 2) +
  geom_smooth(method = "loess", se = FALSE, colour = course_charcoal, linewidth = 1.2) +
  labs(
    title = "A strong relationship with almost zero correlation",
    subtitle = paste0("Simulated correlation = ", round(nonlinear_correlation, 2)),
    x = "X",
    y = "Y = X² + noise"
  )
```

::: {.notes}
Timing: 2 minutes. Correlation captures linear co-movement. Independence implies zero correlation under finite moments, but zero correlation does not imply independence.
:::

## Active check · Conditional or causal?

::: {.exercise}
A wage dataset shows

$$E(\text{wage}\mid\text{degree}=1) - E(\text{wage}\mid\text{degree}=0)=€6.40/\text{hour}.$$

In pairs, write:

1. one sentence this number **does** justify;
2. one causal sentence it **does not yet** justify;
3. two variables that might make the groups unlike.
:::

::: {.notes}
Timing: 3 minutes plus a 1-minute debrief. Candidate confounders: prior ability, family resources, field, labor-market experience, location. Avoid suggesting that adding a long control list automatically identifies the degree effect.
:::

## Pause · 5 minutes {.section-slide background-color="#E85D4A"}

Final block: why averages become reliable.

::: {.notes}
Clock: 02:00–02:05.
:::

## Part V · From samples to populations {.section-slide background-color="#E85D4A"}

The estimate changes when the sample changes.

::: {.notes}
Clock: 02:05–02:25; reserve 02:25–02:30 for the exit ticket and synthesis.
:::

## Population, sample, estimator, estimate

::: columns
::: {.column width="48%"}

### Population target

$$\mu_Y=E(Y)$$

Fixed but usually unknown.

:::
::: {.column width="48%"}

### Sample statistic

$$\bar Y=\frac{1}{n}\sum_{i=1}^{n}Y_i$$

Random before sampling; a number after the data arrive.

:::
:::

::: {.definition .fragment}
An **estimator** is the rule ($\bar Y$). An **estimate** is the realized value from this sample.
:::

::: {.notes}
Timing: 2 minutes. This distinction is foundational for standard errors and confidence intervals.
:::

## Simple random sampling and i.i.d.

We often begin with

$$Y_1,\ldots,Y_n\ \text{independent and identically distributed}.$$

::: {.incremental}

- **Identical distribution:** each observation comes from the same population distribution.
- **Independent:** learning one observation gives no information about another.

:::

::: {.small-note .fragment}
Time series, clustered students, households, firms, and spatial data often violate simple independence. Later methods adapt the uncertainty calculation.
:::

::: {.notes}
Timing: 2 minutes. Say explicitly that i.i.d. is a starting model, not a universal property of data.
:::

## The sample mean has a distribution

Imagine repeatedly drawing samples of $n$ students and recording average commute time.

::: {.incremental}

- Each sample contains different students.
- Each sample produces a different $\bar Y$.
- The distribution of those possible averages is the **sampling distribution**.

:::

::: {.key-idea .fragment}
A histogram of the observed $Y_i$ values is not the same object as the sampling distribution of $\bar Y$.
:::

::: {.notes}
Timing: 2 minutes. The population distribution describes individuals; the sampling distribution describes an estimator across repeated samples.
:::

## What averaging buys us

If $Y_1,\ldots,Y_n$ are i.i.d. with mean $\mu_Y$ and variance $\sigma_Y^2$, then

$$
E(\bar Y)=\mu_Y,
\qquad
\operatorname{Var}(\bar Y)=\frac{\sigma_Y^2}{n},
\qquad
\operatorname{SD}(\bar Y)=\frac{\sigma_Y}{\sqrt n}.
$$

::: {.key-idea .fragment}
Four times the sample size gives half the standard deviation—not one quarter.
:::

::: {.notes}
Timing: 3 minutes. The first equality is unbiasedness. The last quantity motivates the standard error once sigma is estimated.
:::

## Law of large numbers · destination

```{r}
#| label: lln-plot
#| echo: false
#| fig-width: 10
#| fig-height: 5.5

lln_draws <- rexp(500, rate = 1)
lln_data <- data.frame(
  n = seq_along(lln_draws),
  running_mean = cumsum(lln_draws) / seq_along(lln_draws)
)

ggplot(lln_data, aes(n, running_mean)) +
  geom_hline(yintercept = 1, colour = course_red, linewidth = 1.2) +
  geom_line(colour = course_charcoal, linewidth = 0.9) +
  annotate("text", x = 440, y = 1.10, label = "Population mean = 1", colour = course_red_dark, size = 5.5) +
  labs(
    title = "As the sample grows, the running average settles",
    subtitle = "One simulated sequence from a right-skewed exponential population",
    x = "Sample size, n",
    y = "Running sample mean"
  )
```

::: {.notes}
Timing: 2 minutes. One path illustrates, but does not prove, the LLN. State the result: the probability of being far from mu becomes small as n grows under suitable conditions.
:::

## Central limit theorem · shape

For large $n$,

$$
\frac{\bar Y-\mu_Y}{\sigma_Y/\sqrt n}
\overset{d}{\approx}N(0,1).
$$

::: {.incremental}

- The population itself need not be normal.
- A skewed or discrete population can produce nearly normal averages.
- “Large enough” depends on skewness, tails, and dependence.

:::

::: {.notes}
Timing: 2 minutes. Keep convergence notation informal. The standardized average has a stable scale; the unstandardized sampling distribution collapses toward mu.
:::

## Watching the CLT work

```{r}
#| label: clt-simulation
#| echo: false
#| fig-width: 11
#| fig-height: 5.6

sample_sizes <- c(1, 5, 30)
repetitions <- 5000

clt_data <- do.call(rbind, lapply(sample_sizes, function(sample_size) {
  sample_means <- replicate(
    repetitions,
    mean(rexp(sample_size, rate = 1))
  )
  data.frame(
    z = sqrt(sample_size) * (sample_means - 1),
    sample_size = factor(
      paste0("n = ", sample_size),
      levels = paste0("n = ", sample_sizes)
    )
  )
}))

normal_reference <- expand.grid(
  z = seq(-4, 5, length.out = 500),
  sample_size = factor(
    paste0("n = ", sample_sizes),
    levels = paste0("n = ", sample_sizes)
  )
)
normal_reference$density <- dnorm(normal_reference$z)

ggplot(clt_data, aes(z)) +
  geom_density(fill = course_pale, colour = course_red_dark, linewidth = 0.9) +
  geom_line(
    data = normal_reference,
    aes(z, density),
    colour = course_charcoal,
    linewidth = 1,
    inherit.aes = FALSE
  ) +
  facet_wrap(~sample_size, nrow = 1) +
  coord_cartesian(xlim = c(-4, 5), ylim = c(0, 0.48)) +
  labs(
    title = "Averages of skewed observations become approximately normal",
    subtitle = "Red density: simulated standardized averages · Black curve: N(0,1)",
    x = "Standardized sample average",
    y = "Density"
  )
```

::: {.notes}
Timing: 3 minutes. At n=1 the standardized variable retains exponential skewness. By n=30 the approximation is useful, though not magically exact.
:::

## LLN and CLT answer different questions

| Result | What it tells us | Mental image |
|---|---|---|
| Law of large numbers | $\bar Y$ gets close to $\mu_Y$ | the target becomes accurate |
| Central limit theorem | the standardized error becomes approximately normal | the error gets a useful shape |

::: {.key-idea .fragment}
LLN supports consistency. CLT supports approximate probabilities, confidence intervals, and tests.
:::

::: {.notes}
Timing: 2 minutes. This distinction is often blurred. The LLN is about concentration; the CLT is about distributional approximation.
:::

## Quick calculation · the square-root rule

::: {.exercise}
Individual monthly earnings have a standard deviation of €1,200.

1. What is the standard deviation of the sample mean for $n=100$ i.i.d. workers?
2. What sample size cuts that value in half?
3. What assumption would be doubtful if many workers belong to the same household or firm?
:::

::: {.answer .fragment}
1. €120. 2. $n=400$. 3. Independence; shared shocks create within-group dependence.
:::

::: {.notes}
Timing: 3 minutes. Call the sampling SD a standard error once linked to an estimator. Leave finite-population corrections for later.
:::

## One complete econometric story

::: {.incremental}

**Question:** How much would a cigarette tax reduce smoking?

**Estimand:** the average causal response of cigarette consumption to a tax-induced price change.

**Data:** state-by-year prices, taxes, sales, and incomes—a panel.

**Threat:** smoking preferences can affect both political tax choices and consumption.

**Methodological need:** a design that creates credible price variation, plus uncertainty that respects the panel structure.

:::

::: {.notes}
Timing: 2 minutes. This returns to Chapter 1 and previews simultaneous causality and instrumental variables without teaching them now.
:::

## Exit ticket · 3 minutes

::: {.exercise}
Choose one economic question and write four lines:

1. **Goal:** description, prediction, or causation?
2. **Target:** what population quantity would answer it?
3. **Data:** what entities and times would you observe?
4. **Threat:** what could make your comparison misleading?
:::

::: {.notes}
Collect or discuss two answers. Use these to diagnose whether students distinguish a variable from an estimand and an association from a causal effect.
:::

## Takeaways {.section-slide background-color="#E85D4A"}

Econometrics makes the chain visible:

**question → estimand → data → assumptions → evidence → decision**

::: {.notes}
Timing: 2 minutes. Revisit the opening student question and walk it through the chain in one minute.
:::

## What to retain for Lecture 2

::: {.incremental}

- Conditional means describe systematic differences across groups.
- Sample estimates vary; standard errors describe that sampling variation.
- Large samples help, but they do not repair a bad comparison or bad measurement.
- A causal claim needs a design, not only a coefficient.

:::

::: {.small-note}
Primary course source: Stock and Watson, *Introduction to Econometrics*, Chapters 1–2 (materials supplied in `book/chapter-1-2/`). Additional examples and simulations were created for this lecture.
:::

::: {.notes}
Optional final minute. Preview the next lecture’s move from conditional means to estimation and uncertainty.
:::
