---
title: "Lecture XX: A Clear Econometric Question"
subtitle: "Econometric Theory and Practice I"
author: "Swapnil Singh"
date: "2026-09-01"
date-format: "D MMMM YYYY"
---

```{r}
#| label: setup
#| include: false

library(ggplot2)

set.seed(2026)

course_red <- "#E85D4A"
course_charcoal <- "#242424"

theme_set(
  theme_minimal(base_size = 18) +
    theme(
      plot.title = element_text(
        colour = course_charcoal,
        face = "bold"
      ),
      panel.grid.minor = element_blank(),
      plot.background = element_rect(
        fill = "transparent",
        colour = NA
      )
    )
)
```

## Today

::: {.key-idea}
**Central question:** What do we learn from the conditional relationship between an outcome and a regressor?
:::

::: {.incremental}

- Translate an economic question into an estimand.
- Inspect the data before estimating a model.
- Interpret a regression coefficient in context.
- Separate statistical evidence from an economic claim.

:::

::: {.notes}
Use this slide to state the destination before introducing notation. The list is incremental so each goal can be discussed separately.
:::

## From a question to an estimand

::: columns
::: {.column width="48%"}

### Economic question

How does fuel efficiency vary with vehicle weight, holding horsepower fixed?

:::

::: {.column width="48%"}

### Statistical object

$$
E[\text{mpg}_i \mid \text{wt}_i, \text{hp}_i]
= \beta_0 + \beta_1\text{wt}_i + \beta_2\text{hp}_i.
$$

:::
:::

::: {.definition .fragment}
The coefficient $\beta_1$ is a **conditional association**. A causal interpretation requires additional assumptions and a credible design.
:::

## Build the model one line at a time

```{r}
#| label: fit-model
#| echo: true
#| code-line-numbers: "1|2|3"
#| output-location: fragment

model <- lm(mpg ~ wt + hp, data = mtcars)
estimates <- coef(summary(model))
round(estimates, 3)
```

::: {.small-note}
The highlighted lines advance with the slide. Press <kbd>Space</kbd> or the right arrow to move through them.
:::

## Code beside its output

```{r}
#| label: plot-data
#| echo: true
#| output-location: column-fragment
#| fig-width: 7
#| fig-height: 5

ggplot(mtcars, aes(wt, mpg)) +
  geom_point(
    colour = course_red,
    size = 3,
    alpha = 0.82
  ) +
  geom_smooth(
    method = "lm",
    se = FALSE,
    colour = course_charcoal,
    linewidth = 1.2
  ) +
  labs(
    title = "Fuel efficiency falls as weight rises",
    x = "Weight (1,000 lb)",
    y = "Miles per gallon"
  )
```

## A compact regression table

```{r}
#| label: coefficient-table
#| echo: false
#| tbl-cap: "OLS estimates using the built-in mtcars data"

coefficient_table <- broom::tidy(model, conf.int = TRUE)
coefficient_table <- subset(
  coefficient_table,
  term != "(Intercept)",
  select = c(term, estimate, std.error, conf.low, conf.high)
)
names(coefficient_table) <- c(
  "Term", "Estimate", "Std. error", "CI low", "CI high"
)
coefficient_table[-1] <- lapply(coefficient_table[-1], round, 3)

knitr::kable(coefficient_table, align = c("l", "r", "r", "r", "r"))
```

::: {.notes}
Ask students to interpret the weight estimate in the data's original units before discussing uncertainty.
:::

## Check the interpretation

::: {.exercise}
**Think–pair–share (2 minutes).** Write one sentence interpreting the coefficient on `wt`. State the comparison, units, and what is held fixed.
:::

::: {.answer .fragment}
Among cars with the same horsepower, an additional 1,000 lb of weight is associated with the estimated change in miles per gallon shown in the table.
:::

## Useful modes while presenting

::: columns
::: {.column width="48%"}

- <kbd>S</kbd>: speaker view and notes
- <kbd>O</kbd>: slide overview
- <kbd>F</kbd>: full screen
- <kbd>B</kbd>: pause with a black screen

:::

::: {.column width="48%"}

- Use fragments to reveal an argument.
- Use line highlighting to walk through R.
- Use the chalkboard for live annotation.
- Use `.scrollable` only for unavoidable overflow.

:::
:::

## Takeaways {.section-slide background-color="#E85D4A"}

Good econometrics slides make the chain visible:

**question → estimand → code → evidence → interpretation**

::: {.notes}
End by returning to the central question and asking what changed between the unconditional plot and the conditional regression.
:::
